---
title: "Prepare Grunnkart data"
folderName: code # Leave to be filled out by ecRxiv during typesetting (name of the folder under 'indicators/'). 
# If it's a single indicator being reported, the folder name should be the same as the indicatorID. 
# If there are multiple ID's then the folder name could look something like XX_YYY_ZZZ-ÅÅÅ.
country: Norway
continent: Europe
yearAdded: 2026 # the year when the indicator was first submitted to ecRxiv
yearLastUpdate: 2026
status: incomplete # options are complete, incomplete, or deprecated
Version: "0.1.0" # See wiki page: https://github.com/NINAnor/ecRxiv/wiki#versioning
VersionComment: 'First version. The code has been adapted from https://github.com/NINAnor/ecRxiv/tree/main/indicators/NO_URGR_001-005. Template adapated from Anders Kolstad ecRxiv repository: https://github.com/NINAnor/ecRxiv.'
url: https://github.com/NINAnor/egd_data_preparation
# To be filled out by ecRxiv during the typesetting, when they will replace XX_YYYY_ZZ with the name of the root folder for the current indicator(s). It should be the same as the indicatorID, but if there are more indicator IDs presented in the same folder, the folder name might be different, e.g. XX_YYYY_AAA-BBB 
format: 
  html:
    embed-resources: true
    code-fold: true
    toc: true
    toc-title: Contents
    toc-depth: 3
    smooth-scroll: true
execute: 
  cache: true
author:
  - name: Sylvie Clappe             # Enter name
    email: sylvie.clappe@nina.no  # Enter email
    affiliations:
      - id: myID
        name: Norwegian Institute for Nature Research # Enter affiliations
  - name: Bálint Czúcz              
    email: balint.czucz@nina.no             #  Enter subsequent authors like this, or remove if not relevant
    affiliations:
      - ref: myID               # To reuse affiliations reference the id like this
AuthorList: Clappe, S., Czúcz, B.
date: July 13, 2026
bibliography: references.bib
callout-icon: false
lightbox: true
css: ../../../style.css
code-links:
      - text: Add a review
        icon: github
        href: https://github.com/NINAnor/ecRxiv
hide: TRUE # Change to FALSE
---

<!--# This is a template for how to document the indicator analyses. Make sure also to not change the order, or modify, the headers, unless you really need to. This is because it easier to read if all the indicators are presented using the same layout. If there is one header where you don't have anything to write, just leave the header as is, and don't write anything below it. If you are providing code, be careful to annotate and comment on every step in the analysis. Before starting it is recommended to fill in as much as you can in the metadata file. This file will populate the initial table in your output.-->

<!--# Load all you dependencies here -->

```{r setup}
#| include: false
#| cache: false
library(knitr)
library(tidyverse)
library(kableExtra)
library(here)
library(yaml)
library(tibble)
library(conflicted)
library(tidyverse)
library(dplyr)
library(terra)
library(carrier)
library(mirai)

# weird workarounds for cache issues
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::lag)
conflicts_prefer(dplyr::select)
pull<-dplyr::pull

knitr::opts_chunk$set(echo = TRUE)
```

<!--# The following parts are autogenerated. Do not edit. -->


```{r source}
#| echo: false
#| cache: false

source(here::here("_common.R"))
```


```{r reviewers}
#| echo: false
#| results: asis

if (!is.null(yaml_data$codeReviewers)) {
  reviewers <- purrr::map_chr(yaml_data$codeReviewers, "name")
  
  htmltools::HTML(paste0(
    '<div id="title-block-header" class="quarto-title-block default">
       <div class="quarto-title-meta">
         <div>
           <div class="quarto-title-meta-heading">Reviewed by</div>
           <div class="quarto-title-meta-contents">',
    paste(reviewers, collapse = "; "),
    '</div>
         </div>
       </div>
     </div>',
    sep = ""
  ))
}
```

> **Recommended citation**: `r paste(auth, " ", year, ". ", "v. ", version, ": ", url, sep="")`



::: {.callout-tip collapse="true"}
## Metadata
```{r tbl-meta2}
#| tbl-cap: 'Indicator metadata'
#| echo: false
#| warning: false
meta |>
  dplyr::filter(Variable %in% c(
    "Indicator ID",
    "Indicator Name",
    "verbatimName",
    "Continent",
    "Country",
    "Year added", 
    "Last update",
    "Version",
    "Version comment")) |>   
  kable()

```
:::

::: {.callout-tip collapse="true"}

## Log

<!--# Update this logg with short messages for each update -->
- 13 July 2026 - First draft version
:::


<hr />

<!--# Document you work below.  -->

## 1. Summary
This code shows how to download and pre-process Grunnkart data to compile the URGR URAQ and TCCD indicators from ecRxiv: https://github.com/NINAnor/ecRxiv/tree/main/indicators/NO_URGR_001-005.


## 2. Data characteristics

Grunnkart is Norway's official *ecosystem type map* designed to support ecosystem accounting (@strand2024; @gk2). It is produced by NIBIO and SSB.

### 2.1 Spatial and temporal resolution and extent
Spatial resolution is not determined in the metadata, so is temporal resolution. The data are a composite of several datasets from different spatial and temporal resolution. For the purpose of our work, we will assume that the data are from 2025 as it is the date of its release.


| **Grunnkart**  | No, only upon request to Grunnkart is produced by [NIBIO](https://kartkatalog.geonorge.no/metadata/nasjonalt-grunnkart-for-arealanalyse-testversjon-2/28c28e3a-d88f-4a34-8c60-5efe6d56a44d) | We used testversion 0.2, description available in @strand2024 and @gk2 |



### 2.2 Sampling design
The data are a composite of several datasets that have been compiled together. See NIBIO report @strand2024 for more information.

### 2.3 Original units
Original units are polygons.


### 2.4 Instructions for citing, using and accessing data
Data are available on Geonorge [here]https://kartkatalog.geonorge.no/metadata/nasjonalt-grunnkart-for-arealanalyse-testversjon-2/28c28e3a-d88f-4a34-8c60-5efe6d56a44d). Depending on the user's status, these data can be either freely available or available upon request only.


### 2.5 Additional comments about the dataset
Other metadata can be found [on Geonorge](https://kartkatalog.geonorge.no/metadata/nasjonalt-grunnkart-for-arealanalyse/28c28e3a-d88f-4a34-8c60-5efe6d56a44d?search=grunnkart)


## 3. Uncertainties
### 3.1 General uncertainties
These data have various sources of uncertainties but they could not be quantified and reported. These sources have not been discussed in the reports associated with the data but they are linked, among other things, to: the uncertainty of the different data sources used to compile the Grunnkart, the hierarchical choices made during the classification of ecosystem types, the crosswalk used between all the classifications used in the different datasets used to produce the Grunnkart.

### 3.2 Reported uncertainties
Uncertainties related to this dataset were not quantified or reported by NIBIO.


## 4. Data download
The Grunnkart is a a big dataset and, for NINA, these data were only available upon request. This means that we could not read them using an API and they had to be downloaded and stored on NINA's internal servers directly.

<!-- ```{r} -->
<!-- #| eval: false -->
<!-- #| code-summary: "Read the .tif from a zip file" -->

<!-- # Call functions -->
<!-- source(here::here("src", "functions", "list_zip_files.R")) -->
<!-- source(here::here("src", "functions", "extract_zip_files.R")) -->

<!-- # Path to the downloaded Copernicus data -->
<!-- path_imd <- "/data/P-Prosjekter2/412413_2023_no_egd/git_data/IMD_Copernicus" -->

<!-- # List datasets' paths to read and their names -->
<!-- files_imd_read <- files_list_sev_fn(path_imd) -->

<!-- # List datasets path (here we remove the zip marker and the name of the file) -->
<!-- files_imd_write <- str_remove(files_imd_read[[1]], "CLMS.*") %>% -->
<!--                       str_remove("/vsizip/\\{") -->

<!-- # Run the function to read and write the rasters from the zip files directly -->
<!-- library(mirai) -->
<!-- daemons(10) -->
<!-- map(seq_along(files_imd_read[[1]]), -->
<!--     in_parallel( ~process_raster_fn(path_read = files_imd_read[[1]][.x], -->
<!--                                     path_write = files_imd_write[.x], -->
<!--                                     files_names = files_imd_read[[2]][.x]), -->
<!--                  process_raster_fn = process_raster_fn, -->
<!--                  files_imd_read = files_imd_read, -->
<!--                  files_imd_write = files_imd_write -->
<!--     ) -->
<!-- ) -->

<!-- daemons(0) -->

<!-- ``` -->

<!-- ### 4.2 Copernicus IMD Confidence layers -->
<!-- Now we do the same for the confidence layers. -->
<!-- ```{r} -->
<!-- #| eval: false -->
<!-- #| code-summary: "Read the .tif from a zip file" -->

<!-- # Call functions -->
<!-- source(here::here("src", "functions", "list_zip_files.R")) -->
<!-- source(here::here("src", "functions", "extract_zip_files.R")) -->

<!-- # Path to the downloaded Copernicus data -->
<!-- path_imd_unc <- "/data/P-Prosjekter2/412413_2023_no_egd/git_data/IMD_Copernicus_uncertainties" -->

<!-- # List datasets' paths to read and their names -->
<!-- files_imd_read <- files_list_sev_fn(path_imd_unc) -->

<!-- # List datasets path (here we remove the zip marker and the name of the file) -->
<!-- files_imd_write <- str_remove(files_imd_read[[1]], "CLMS.*") %>% -->
<!--                       str_remove("/vsizip/\\{") -->

<!-- # Run the function to read and write the rasters from the zip files directly -->
<!-- library(mirai) -->
<!-- daemons(10) -->
<!-- map(seq_along(files_imd_read[[1]]), -->
<!--     in_parallel( ~process_raster_fn(path_read = files_imd_read[[1]][.x], -->
<!--                                     path_write = files_imd_write[.x], -->
<!--                                     files_names = files_imd_read[[2]][.x]), -->
<!--                  process_raster_fn = process_raster_fn, -->
<!--                  files_imd_read = files_imd_read, -->
<!--                  files_imd_write = files_imd_write -->
<!--     ) -->
<!-- ) -->

<!-- daemons(0) -->

<!-- ``` -->



<!-- ## 5. Data cleaning and formatting -->
<!-- Once the rasters (.tf) have been extracted and written on the disk, they can be read to create a virtual raster that will be used for the subsequent analyses later on. -->

<!-- ### 5.1 Copernicus IMD -->

<!-- ```{r} -->
<!-- #| eval: false -->
<!-- #| code-summary: "Create Copernicus IMD virtual raster" -->

<!-- # Code co-created by Sylvie Clappe and Balint Czucz -->

<!-- # Create or load the IMD virtual rasters -->
<!-- terra::setGDALconfig("GTIFF_SRS_SOURCE", "EPSG") # to quiet a GDAL warning -->
<!-- path_imd_list <- dir(path_imd, recursive = FALSE, full.names = TRUE) -->
<!-- yrs_imd <- str_sub(path_imd_list, start = -4) -->
<!-- file_imd <- map_vec(seq_along(path_imd_list), function(i){paste0(path_imd_list[i], "/IMD_", yrs_imd[i], ".vrt")}) -->

<!-- imd_vrt <- map(seq_along(file_imd), ~  -->

<!--                 # if it does not exist yet, create the virtual raster -->
<!--                if (!file.exists(file_imd[.x])) { -->
<!--                     path_imd_list[.x] %>% -->
<!--                     list.files(pattern= '.tif$', all.files= TRUE, full.names= TRUE) %>% # list of tiles -->
<!--                     vrt(filename= file_imd[.x])  -->

<!--                 # otherwise just load it  -->
<!--                } else { -->
<!--                 rast(file_imd[.x])   -->
<!--                } -->

<!--               ) -->

<!-- ``` -->


<!-- ### 5.2 Copernicus IMD Confidence layers -->
<!-- ```{r} -->
<!-- #| eval: false -->
<!-- #| code-summary: "Create Copernicus IMD confidence virtual raster" -->

<!-- # Code co-created by Sylvie Clappe and Balint Czucz -->

<!-- # Create or load the IMD virtual rasters -->
<!-- terra::setGDALconfig("GTIFF_SRS_SOURCE", "EPSG") # to quiet a GDAL warning -->
<!-- path_imd_list <- dir(path_imd_unc, recursive = FALSE, full.names = TRUE) -->
<!-- yrs_imd <- str_sub(path_imd_list, start = -4) -->
<!-- file_imd <- map_vec(seq_along(path_imd_list), function(i){paste0(path_imd_list[i], "/IMD_", yrs_imd[i], ".vrt")}) -->

<!-- imd_vrt <- map(seq_along(file_imd), ~  -->

<!--                 # if it does not exist yet, create the virtual raster -->
<!--                if (!file.exists(file_imd[.x])) { -->
<!--                     path_imd_list[.x] %>% -->
<!--                     list.files(pattern= '.tif$', all.files= TRUE, full.names= TRUE) %>% # list of tiles -->
<!--                     vrt(filename= file_imd[.x])  -->

<!--                 # otherwise just load it  -->
<!--                } else { -->
<!--                 rast(file_imd[.x])   -->
<!--                } -->

<!--               ) -->

<!-- ``` -->


## 6. Export file
Virtual raster already exported in the previous code.

```{r export}
#| eval: false
```

## 7. References
<!--# You can add references manually or use a citation manager and add intext citations as with crossreferencing and hyperlinks. See https://quarto.org/docs/authoring/footnotes-and-citations.html -->


::: {#refs}
:::

## 8. Author contributions

<!-- Fill in CReDIT and MeRIT statments to clarify the contributions from individual authors. For MeRIT, see https://www.nature.com/articles/s41467-023-37039-1. For CRedit, see https://credit.niso.org/. You can start with the template structur below, and just delete the rows you dont use.  -->

**CRediT statement**

* Conceptualization: SC
* Methodology: SC & BC
* Data curation & formatting: SC
* Writing – original draft: SC
* Writing – review & editing: BC
* Document structure: SC & AK


**MeRIT attribution**

The structure of the document has been taken from ecRxiv's template for condition indicators, created by Anders Kolstad (AK). SC adapted it to the purpose of the project.

**Study design**
The study was designed by SC and BC.

**Data processing**
Raw data were downloaded by SC. They were curated and cleaned by SC and BC. Code was written by both SC and BC.


::: {.callout-tip collapse="true"}
## Session Info
<!-- You can leave this last part as it is, unless you have a good reason to alter it. -->

This information is mainly aimed at developers that need to recreate the R environment that produced the html.
Important note: this information is about the R session that was running when the html was rendered, and if some of the code chunks were not evaluated when the html was rendered, they may have been executed in a different environment.  

```{r sessionInfo}
#| echo: false
sessionInfo()
```

:::
