---
title: "Prepare Grunnkart data"
folderName: code # Leave to be filled out by ecRxiv during typesetting (name of the folder under 'indicators/'). 
# If it's a single indicator being reported, the folder name should be the same as the indicatorID. 
# If there are multiple ID's then the folder name could look something like XX_YYY_ZZZ-ÅÅÅ.
country: Norway
continent: Europe
yearAdded: 2026 # the year when the indicator was first submitted to ecRxiv
yearLastUpdate: 2026
status: incomplete # options are complete, incomplete, or deprecated
Version: "0.1.0" # See wiki page: https://github.com/NINAnor/ecRxiv/wiki#versioning
VersionComment: 'First version. The code has been adapted from https://github.com/NINAnor/ecRxiv/tree/main/indicators/NO_URGR_001-005. Template adapated from Anders Kolstad ecRxiv repository: https://github.com/NINAnor/ecRxiv.'
url: https://github.com/NINAnor/egd_data_preparation
# To be filled out by ecRxiv during the typesetting, when they will replace XX_YYYY_ZZ with the name of the root folder for the current indicator(s). It should be the same as the indicatorID, but if there are more indicator IDs presented in the same folder, the folder name might be different, e.g. XX_YYYY_AAA-BBB 
format: 
  html:
    embed-resources: true
    code-fold: true
    toc: true
    toc-title: Contents
    toc-depth: 3
    smooth-scroll: true
execute: 
  cache: true
author:
  - name: Sylvie Clappe             # Enter name
    email: sylvie.clappe@nina.no  # Enter email
    affiliations:
      - id: myID
        name: Norwegian Institute for Nature Research # Enter affiliations
  - name: Bálint Czúcz              
    email: balint.czucz@nina.no             #  Enter subsequent authors like this, or remove if not relevant
    affiliations:
      - ref: myID               # To reuse affiliations reference the id like this
AuthorList: Clappe, S., Czúcz, B.
date: July 13, 2026
bibliography: references.bib
callout-icon: false
lightbox: true
css: ../../../style.css
code-links:
      - text: Add a review
        icon: github
        href: https://github.com/NINAnor/ecRxiv
hide: TRUE # Change to FALSE
---

<!--# This is a template for how to document the indicator analyses. Make sure also to not change the order, or modify, the headers, unless you really need to. This is because it easier to read if all the indicators are presented using the same layout. If there is one header where you don't have anything to write, just leave the header as is, and don't write anything below it. If you are providing code, be careful to annotate and comment on every step in the analysis. Before starting it is recommended to fill in as much as you can in the metadata file. This file will populate the initial table in your output.-->

<!--# Load all you dependencies here -->

```{r setup}
#| include: false
#| cache: false
library(knitr)
library(tidyverse)
library(kableExtra)
library(here)
library(yaml)
library(tibble)
library(conflicted)
library(tidyverse)
library(dplyr)
library(sf)
library(carrier)
library(mirai)

# weird workarounds for cache issues
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::lag)
conflicts_prefer(dplyr::select)
pull<-dplyr::pull

knitr::opts_chunk$set(echo = TRUE)
```

<!--# The following parts are autogenerated. Do not edit. -->


```{r source}
#| echo: false
#| cache: false

source(here::here("_common.R"))
```


```{r reviewers}
#| echo: false
#| results: asis

if (!is.null(yaml_data$codeReviewers)) {
  reviewers <- purrr::map_chr(yaml_data$codeReviewers, "name")
  
  htmltools::HTML(paste0(
    '<div id="title-block-header" class="quarto-title-block default">
       <div class="quarto-title-meta">
         <div>
           <div class="quarto-title-meta-heading">Reviewed by</div>
           <div class="quarto-title-meta-contents">',
    paste(reviewers, collapse = "; "),
    '</div>
         </div>
       </div>
     </div>',
    sep = ""
  ))
}
```

> **Recommended citation**: `r paste(auth, " ", year, ". ", "v. ", version, ": ", url, sep="")`



::: {.callout-tip collapse="true"}
## Metadata
```{r tbl-meta2}
#| tbl-cap: 'Indicator metadata'
#| echo: false
#| warning: false
meta |>
  dplyr::filter(Variable %in% c(
    "Indicator ID",
    "Indicator Name",
    "verbatimName",
    "Continent",
    "Country",
    "Year added", 
    "Last update",
    "Version",
    "Version comment")) |>   
  kable()

```
:::

::: {.callout-tip collapse="true"}

## Log

<!--# Update this logg with short messages for each update -->
- 13 July 2026 - First draft version
:::


<hr />

<!--# Document you work below.  -->

## 1. Summary
This code shows how to download and pre-process Grunnkart data to compile the URGR URAQ and TCCD indicators from ecRxiv: https://github.com/NINAnor/ecRxiv/tree/main/indicators/NO_URGR_001-005.


## 2. Data characteristics
Grunnkart is Norway's official *ecosystem type map* designed to support ecosystem accounting (@strand2024; @gk2). It is produced by NIBIO and SSB.

### 2.1 Spatial and temporal resolution and extent
Spatial resolution is not determined in the metadata, so is temporal resolution. The data are a composite of several datasets from different spatial and temporal resolution. For the purpose of our work, we will assume that the data are from 2025 as it is the date of its release.


### 2.2 Sampling design
The data are a composite of several datasets that have been compiled together. See NIBIO report @strand2024 for more information.

### 2.3 Original units
Original units are polygons.


### 2.4 Instructions for citing, using and accessing data
Data are available on Geonorge [here]https://kartkatalog.geonorge.no/metadata/nasjonalt-grunnkart-for-arealanalyse-testversjon-2/28c28e3a-d88f-4a34-8c60-5efe6d56a44d). Depending on the user's status, these data can be either freely available or available upon request only.


### 2.5 Additional comments about the dataset
Other metadata can be found [on Geonorge](https://kartkatalog.geonorge.no/metadata/nasjonalt-grunnkart-for-arealanalyse/28c28e3a-d88f-4a34-8c60-5efe6d56a44d?search=grunnkart)


## 3. Uncertainties
### 3.1 General uncertainties
These data have various sources of uncertainties but they could not be quantified and reported. These sources have not been discussed in the reports associated with the data but they are linked, among other things, to: the uncertainty of the different data sources used to compile the Grunnkart, the hierarchical choices made during the classification of ecosystem types, the crosswalk used between all the classifications used in the different datasets used to produce the Grunnkart.

### 3.2 Reported uncertainties
Uncertainties related to this dataset were not quantified or reported by NIBIO.


## 4. Load the data
The Grunnkart is a a big dataset and, for NINA, these data were only available upon request. This means that we could not read them using an API and they had to be downloaded and stored on NINA's internal servers directly. 

From experience, there are many geometrical issues with the Grunnkart that need to be fixed. In order to do that properly, we need to ensure that a copy of the raw data remains somewhere, untouched. That is why the loading of the data is just about loading the orginial version of the data, without doing any pre-cleaning, including re-projection of the tiles that are not in EPSG:25832. This strategy though puts a strain on the memory, that is why we then save the raw data as a geopackage file that we can read in later on. It is necessary to keep the raw data as the transformations that will be required to clean the data may create empty geometries despite the geometry being present in the raw data. So it is important to keep the raw data to help correct these issues. 

*This step can be skipped if the .gpkg has already been created*.

```{r}
#| eval: false
#| code-summary: "Read the sf objects from gdb"
# Code co-created by Sylvie Clappe and Jennifer Hansen

# Call a function to read gdb files
source(here::here("src", "functions", "read_gdb_files.R"))

# Path to the Grunnkart  
gdb_folder <- "/data/R/GeoSpatialData/LandUse/Norway_Arealregnskap/Original/GrunnkartArealregnskap FGDB-format"

# List the gdb files
gdb_files <- list.files(gdb_folder, pattern = "_gdb\\.gdb$", 
                        full.names = TRUE)

# Read data with gdb files
gk_rw_data <- gdb_files %>% 
                 map(\(.) read_gdb(.)) 

# Name the layers
names(gk_rw_data) <- str_extract(gdb_files, "(?<=format/).{2}")

# Save an RDS copy on scratch
# Path to the result geopackage
gpkg_path <- paste0(str_remove(tempdir(), "Rtmp\\w*"), "GK_raw/GK_raw_layer.gpkg")

iwalk(gk_rw_data, function(layer, name) {
        st_write(obj = layer,
                 dsn = gpkg_path,
                 layer = name,   
                 append = FALSE,    
                 delete_layer = TRUE)})


# Clean R environment
rm(list=setdiff(ls(), c("pull",
                        "gpkg_path",
                        "gdb_files")))

```

Now that we the raw data, we can move on to the cleaning and formatting step.


## 5. Data cleaning and formatting
*Note: most of the code could not be run in parallel due to the size of the data. `in_parallel()` copies the data to each deamon used, while `future_map()` is not advised when the data are too big as moving them to the GlobalEnv once the operation is finished will take a lot of time, ruining the time you gain by running the operations in parallel.*

First we need to re-project the tiles that are not in EPSG:25832. This CRS was chosen as it is the one most tiles had. We re-project but still keep the tiles separated. We also write a geopackage as well to keep these data to solve issues if any occur during the cleaning process.
```{r}
#| eval: false
#| code-summary: "Re-projection to EPSG:25832"

# Read RDS file with raw data
gk_rw_data <- st_read(gpkg_path)


# Define CRS of the first layer (most common CRS of the gdb files)
gk_crs <- st_crs(gk_rw_data [[1]])


# Re-project if needed
gk_data <- map(gk_rw_data, 
               function(i){
  
  if(st_crs(i) != gk_crs){
  st_transform(i, crs = gk_crs)
    
}else{
  i
}})

# Path to the result geopackage
gpkg_path_proj <- paste0(str_remove(tempdir(), "Rtmp\\w*"), "GK_raw/GK_projected_layer.gpkg")

# Save as gpkg
iwalk(gk_data, function(layer, name) {
        st_write(obj = layer,
                 dsn = gpkg_path_proj,
                 layer = name,   
                 append = FALSE,    
                 delete_layer = TRUE)})

# Read if dataset already available
layer_names <- st_layers(gpkg_path_proj)$name
gk_data <- map(layer_names, ~st_read(gpkg_path_proj, layer = .x, quiet = TRUE))
names(gk_data) <- str_extract(gdb_files, "(?<=format/).{2}")


# Clean R environment
rm(list=setdiff(ls(), c("pull",
                        "gdb_files",
                        "gpkg_path",
                        "gpkg_path_proj")))

```

We then re-calculate the area of the polygons. If the newly calculated area under 0.05 m2, we compare it with the area from the original dataset. We chose the threshold arbitrarily as the Grunnkart doesn't have any MMU. We isolate these polygons and check if they are slivers by randomly inspecting some in QGIS. If they are slivers, we delete them. This needs to be done prior to fixing any geometrical issue as they can cause `st_make_valid()` to fail, split into multi-geometries, or generate anomalous dimensions. We export this intermediate layer so we can revert back to it in case of issues.

```{r}
#| eval: false
#| code-summary: "Clean slivers"

# Calculate areas in the new projection
gk_data <- map(gk_data,
               ~mutate(.x, area_m2_proj = unclass(st_area(.x))))

# Identify polygons with area 0
gk_data_0 <- map(gk_data,
                 ~filter(.x, area_m2_proj =< 0.05)) # Random checks with QGIS
names(gk_data_0) <- str_extract(gdb_files, "(?<=format/).{2}")

# Clean them all after random checkes with QGIS
gk_data <- map(gk_data,
             ~filter(.x, area_m2_proj > 0.05)) 

# Export gk_data without sliver
gpkg_path_sliver <- paste0(str_remove(tempdir(), "Rtmp\\w*"), "GK_raw/GK_proj_sliverscleaned_layer.gpkg")

iwalk(gk_data, function(layer, name) {
        st_write(obj = layer,
                 dsn = gpkg_path_sliver,
                 layer = name,   
                 append = FALSE,    
                 delete_layer = TRUE)})


# Clean R environment
rm(list=setdiff(ls(), c("pull",
                        "gpkg_path",
                        "gpkg_path_proj",
                        "gpkg_path_sliver")))

```


Then we fix the geometry errors. We run `st_make_valid` and if it fails, we revert to `st_buffer(geometry, dist = 0)` which sometimes fixes the issues. As we expect that some empty geometries and geometrical issues will remain, we don't overwrite `gk_data` so we can further investigate if necessary. We will clean the R environment after the geometrical cleaning.
```{r}
#| eval: false
#| code-summary: "Fix geometry errors"

gk_data_valid <- gk_data %>%
                 imap(function(layer, name) {
                      tryCatch({
                        st_make_valid(layer)
                      }, error = function(e) {
                        message("st_make_valid() failed for layer '", name, "': ", conditionMessage(e),
                                "\nFalling back to st_buffer(dist = 0)")
                        st_buffer(layer, dist = 0)
                      })
                    })

# Export the dataset
gpkg_path_valid <- paste0(str_remove(tempdir(), "Rtmp\\w*"), "GK_raw/GK_proj_sliverscln_stvalid_layer.gpkg")

iwalk(gk_data_valid, function(layer, name) {
        st_write(obj = layer,
                 dsn = gpkg_path_valid,
                 layer = name,   
                 append = FALSE,    
                 delete_layer = TRUE)})

# Clean R environment
rm(list=setdiff(ls(), c("pull",
                        "gpkg_path",
                        "gpkg_path_proj",
                        "gpkg_path_sliver",
                        "gpkg_path_valid")))

```


Then we do some checks to see if the corrections were efficient.
```{r}
#| eval: false
#| code-summary: "Check pre-cleaning worked"

valid_check <- lapply(gk_data_valid,
                      function(i){which(st_is_valid(i) == FALSE)})

empty_check <- lapply(gk_data_valid,
                      function(i){which(st_is_empty(i) == TRUE)})

lapply(valid_check, function (i){length(i)})
lapply(empty_check, function (i){length(i)})


```

#I STOPPED HERE###################

Making the polygons valid resulted in a lot of polygons with empty geometry. Also, some polygons were still invalid. We are now looking at these to see how we can further clean them. We start by investigating the empty geometries. There is only one polygon with an empty geometry. Looking at the area, it is a polygon with 20,156.6 m2. It cannot be deleted as it is not a sliver. We will try to retrieve its geometry using the adjacent polygons that we can identify based on the original dataset and the `id` column. (retrieve geometry of a polygon in a wall to wall map based on adjacent polygons R).

how to make polygons valid after already running st_make_valid or st_buffer(sf, 0)

```{r}
#| eval: false
#| code-summary: "Investigate polygons with empty geometry"

# Tile 18
empty_18 <- gk_data_valid$`18`[empty_check$`18`, ]

# Polygon whose geometry is missing
gk_data_valid$`18` <- filter(gk_data_valid$`18`, areal_m2 == 0) 



```




BALINT ADDITIONAL CLEANING STEPS. I didn't do any union before as I thought you might need all the small polygons?


Now we dissolve the ecosystem types into single polygons for further analyses. This should enable us to save some computing time.
```{r}
#| eval: false
#| code-summary: "Dissolve ecosystem types into one single polygon"



```


## 6. Export file

```{r export}
#| eval: false
#| gk_data <- gdb_files %>% 
                 map(\(.) read_gdb(., projection_crs = gk_crs)) %>% 
                 compact() %>%  
                 bind_rows()
```

## 7. References
<!--# You can add references manually or use a citation manager and add intext citations as with crossreferencing and hyperlinks. See https://quarto.org/docs/authoring/footnotes-and-citations.html -->


::: {#refs}
:::

## 8. Author contributions

<!-- Fill in CReDIT and MeRIT statments to clarify the contributions from individual authors. For MeRIT, see https://www.nature.com/articles/s41467-023-37039-1. For CRedit, see https://credit.niso.org/. You can start with the template structur below, and just delete the rows you dont use.  -->

**CRediT statement**

* Conceptualization: SC
* Methodology: SC & BC
* Data curation & formatting: SC
* Writing – original draft: SC
* Writing – review & editing: BC
* Document structure: SC & AK


**MeRIT attribution**

The structure of the document has been taken from ecRxiv's template for condition indicators, created by Anders Kolstad (AK). SC adapted it to the purpose of the project.

**Study design**
The study was designed by SC and BC.

**Data processing**
Raw data were downloaded by SC. They were curated and cleaned by SC and BC. Code was written by both SC and BC.


::: {.callout-tip collapse="true"}
## Session Info
<!-- You can leave this last part as it is, unless you have a good reason to alter it. -->

This information is mainly aimed at developers that need to recreate the R environment that produced the html.
Important note: this information is about the R session that was running when the html was rendered, and if some of the code chunks were not evaluated when the html was rendered, they may have been executed in a different environment.  

```{r sessionInfo}
#| echo: false
sessionInfo()
```

:::
