---
title: "Prepare Grunnkart data"
folderName: code # Leave to be filled out by ecRxiv during typesetting (name of the folder under 'indicators/'). 
# If it's a single indicator being reported, the folder name should be the same as the indicatorID. 
# If there are multiple ID's then the folder name could look something like XX_YYY_ZZZ-ÅÅÅ.
country: Norway
continent: Europe
yearAdded: 2026 # the year when the indicator was first submitted to ecRxiv
yearLastUpdate: 2026
status: incomplete # options are complete, incomplete, or deprecated
Version: "0.1.0" # See wiki page: https://github.com/NINAnor/ecRxiv/wiki#versioning
VersionComment: 'First version. The code has been adapted from https://github.com/NINAnor/ecRxiv/tree/main/indicators/NO_URGR_001-005. Template adapated from Anders Kolstad ecRxiv repository: https://github.com/NINAnor/ecRxiv.'
url: https://github.com/NINAnor/egd_data_preparation
# To be filled out by ecRxiv during the typesetting, when they will replace XX_YYYY_ZZ with the name of the root folder for the current indicator(s). It should be the same as the indicatorID, but if there are more indicator IDs presented in the same folder, the folder name might be different, e.g. XX_YYYY_AAA-BBB 
format: 
  html:
    embed-resources: true
    code-fold: true
    toc: true
    toc-title: Contents
    toc-depth: 3
    smooth-scroll: true
execute: 
  cache: true
author:
  - name: Sylvie Clappe             # Enter name
    email: sylvie.clappe@nina.no  # Enter email
    affiliations:
      - id: myID
        name: Norwegian Institute for Nature Research # Enter affiliations
  - name: Bálint Czúcz              
    email: balint.czucz@nina.no             #  Enter subsequent authors like this, or remove if not relevant
    affiliations:
      - ref: myID               # To reuse affiliations reference the id like this
AuthorList: Clappe, S., Czúcz, B.
date: July 13, 2026
bibliography: references.bib
callout-icon: false
lightbox: true
css: ../../../style.css
code-links:
      - text: Add a review
        icon: github
        href: https://github.com/NINAnor/ecRxiv
hide: TRUE # Change to FALSE
---

<!--# This is a template for how to document the indicator analyses. Make sure also to not change the order, or modify, the headers, unless you really need to. This is because it easier to read if all the indicators are presented using the same layout. If there is one header where you don't have anything to write, just leave the header as is, and don't write anything below it. If you are providing code, be careful to annotate and comment on every step in the analysis. Before starting it is recommended to fill in as much as you can in the metadata file. This file will populate the initial table in your output.-->

<!--# Load all you dependencies here -->

```{r setup}
#| include: false
#| cache: false
library(knitr)
library(tidyverse)
library(kableExtra)
library(here)
library(yaml)
library(tibble)
library(conflicted)
library(tidyverse)
library(dplyr)
library(sf)
library(carrier)
library(mirai)

# weird workarounds for cache issues
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::lag)
conflicts_prefer(dplyr::select)
pull<-dplyr::pull

knitr::opts_chunk$set(echo = TRUE)
```

<!--# The following parts are autogenerated. Do not edit. -->


```{r source}
#| echo: false
#| cache: false

source(here::here("_common.R"))
```


```{r reviewers}
#| echo: false
#| results: asis

if (!is.null(yaml_data$codeReviewers)) {
  reviewers <- purrr::map_chr(yaml_data$codeReviewers, "name")
  
  htmltools::HTML(paste0(
    '<div id="title-block-header" class="quarto-title-block default">
       <div class="quarto-title-meta">
         <div>
           <div class="quarto-title-meta-heading">Reviewed by</div>
           <div class="quarto-title-meta-contents">',
    paste(reviewers, collapse = "; "),
    '</div>
         </div>
       </div>
     </div>',
    sep = ""
  ))
}
```

> **Recommended citation**: `r paste(auth, " ", year, ". ", "v. ", version, ": ", url, sep="")`



::: {.callout-tip collapse="true"}
## Metadata
```{r tbl-meta2}
#| tbl-cap: 'Indicator metadata'
#| echo: false
#| warning: false
meta |>
  dplyr::filter(Variable %in% c(
    "Indicator ID",
    "Indicator Name",
    "verbatimName",
    "Continent",
    "Country",
    "Year added", 
    "Last update",
    "Version",
    "Version comment")) |>   
  kable()

```
:::

::: {.callout-tip collapse="true"}

## Log

<!--# Update this logg with short messages for each update -->
- 13 July 2026 - First draft version
:::


<hr />

<!--# Document you work below.  -->

## 1. Summary
This code shows how to download and pre-process Grunnkart data to compile the URGR URAQ and TCCD indicators from ecRxiv: https://github.com/NINAnor/ecRxiv/tree/main/indicators/NO_URGR_001-005.


## 2. Data characteristics
Grunnkart is Norway's official *ecosystem type map* designed to support ecosystem accounting (@strand2024; @gk2). It is produced by NIBIO and SSB.

### 2.1 Spatial and temporal resolution and extent
Spatial resolution is not determined in the metadata, so is temporal resolution. The data are a composite of several datasets from different spatial and temporal resolution. For the purpose of our work, we will assume that the data are from 2025 as it is the date of its release.


### 2.2 Sampling design
The data are a composite of several datasets that have been compiled together. See NIBIO report @strand2024 for more information.

### 2.3 Original units
Original units are polygons.


### 2.4 Instructions for citing, using and accessing data
Data are available on Geonorge [here]https://kartkatalog.geonorge.no/metadata/nasjonalt-grunnkart-for-arealanalyse-testversjon-2/28c28e3a-d88f-4a34-8c60-5efe6d56a44d). Depending on the user's status, these data can be either freely available or available upon request only.


### 2.5 Additional comments about the dataset
Other metadata can be found [on Geonorge](https://kartkatalog.geonorge.no/metadata/nasjonalt-grunnkart-for-arealanalyse/28c28e3a-d88f-4a34-8c60-5efe6d56a44d?search=grunnkart)


## 3. Uncertainties
### 3.1 General uncertainties
These data have various sources of uncertainties but they could not be quantified and reported. These sources have not been discussed in the reports associated with the data but they are linked, among other things, to: the uncertainty of the different data sources used to compile the Grunnkart, the hierarchical choices made during the classification of ecosystem types, the crosswalk used between all the classifications used in the different datasets used to produce the Grunnkart.

### 3.2 Reported uncertainties
Uncertainties related to this dataset were not quantified or reported by NIBIO.


## 4. Load the data
The Grunnkart is a a big dataset and, for NINA, these data were only available upon request. This means that we could not read them using an API and they had to be downloaded and stored on NINA's internal servers directly. 

From experience, there are many geometrical issues with the Grunnkart that need to be fixed. In order to do that properly, we need to ensure that a copy of the raw data remains somewhere, untouched. That is why the loading of the data is just about loading the orginial version of the data, without doing any pre-cleaning, including re-projection of the tiles that are not in EPSG:25832. This strategy though puts a strain on the memory, that is why we then save the raw data as a geopackage file that we can read in later on. It is necessary to keep the raw data as the transformations that will be required to clean the data may create empty geometries despite the geometry being present in the raw data. So it is important to keep the raw data to help correct these issues. 

*This step can be skipped if the .gpkg has already been created*.

```{r}
#| eval: false
#| code-summary: "Read the sf objects from gdb"
# Code co-created by Sylvie Clappe and Jennifer Hansen

# Call a function to read gdb files
source(here::here("src", "functions", "read_gdb_files.R"))

# Path to the Grunnkart  
gdb_folder <- "/data/R/GeoSpatialData/LandUse/Norway_Arealregnskap/Original/GrunnkartArealregnskap FGDB-format"

# List the gdb files
gdb_files <- list.files(gdb_folder, pattern = "_gdb\\.gdb$", 
                        full.names = TRUE)

# Read data with gdb files
gk_rw_data <- gdb_files %>% 
                 map(\(.) read_gdb(.)) 

# Name the layers
names(gk_rw_data) <- str_extract(gdb_files, "(?<=format/).{2}")

# Export intermediate mayer (if needed)
# gpkg_path <- paste0(str_remove(tempdir(), "Rtmp\\w*"), "GK_raw/GK_raw_layer.gpkg")
# 
# iwalk(gk_rw_data, function(layer, name) {
#         st_write(obj = layer,
#                  dsn = gpkg_path,
#                  layer = name,   
#                  append = FALSE,    
#                  delete_layer = TRUE)})

# Read intermediate layer (if needed)
# gk_rw_data <- st_read(gpkg_path)


# Clean R environment
rm(list=setdiff(ls(), c("pull",
                        "gpkg_path",
                        "gdb_files",
                        "gk_rw_data")))

```

Now that we the raw data, we can move on to the cleaning and formatting step.


## 5. Data cleaning and formatting
*Note: most of the code could not be run in parallel due to the size of the data. `in_parallel()` copies the data to each deamon used, while `future_map()` is not advised when the data are too big as moving them to the GlobalEnv once the operation is finished will take a lot of time, ruining the time you gain by running the operations in parallel.*

First we need to re-project the tiles that are not in EPSG:25832. This CRS was chosen as it is the one most tiles had. We re-project but still keep the tiles separated. We also write a geopackage as well to keep these data to solve issues if any occur during the cleaning process.
```{r}
#| eval: false
#| code-summary: "Re-projection to EPSG:25832"

# Define CRS of the first layer (most common CRS of the gdb files)
gk_crs <- st_crs(gk_rw_data [[1]])


# Re-project if needed
gk_data <- map(gk_rw_data, 
               function(i){
  
  if(st_crs(i) != gk_crs){
  st_transform(i, crs = gk_crs)
    
}else{
  i
}})

# Path to the result geopackage
gpkg_path_proj <- paste0(str_remove(tempdir(), "Rtmp\\w*"), "GK_raw/GK_projected_layer.gpkg")

# Export intermediate layer (if needed)
# iwalk(gk_data, function(layer, name) {
#         st_write(obj = layer,
#                  dsn = gpkg_path_proj,
#                  layer = name,   
#                  append = FALSE,    
#                  delete_layer = TRUE)})

# Read intermediate layer (if needed)
# layer_names <- st_layers(gpkg_path_proj)$name
# gk_data <- map(layer_names, ~st_read(gpkg_path_proj, layer = .x, quiet = TRUE))
# names(gk_data) <- str_extract(gdb_files, "(?<=format/).{2}")


# Clean R environment
rm(list=setdiff(ls(), c("pull",
                        "gdb_files",
                        "gpkg_path",
                        "gpkg_path_proj",
                        "gk_data")))

```

We then re-calculate the area of the polygons. If the newly calculated area under 0.05 m2, we compare it with the area from the original dataset. We chose the threshold arbitrarily as the Grunnkart doesn't have any MMU. We isolate these polygons and check if they are slivers by randomly inspecting some in QGIS. If they are slivers, we delete them. This needs to be done prior to fixing any geometrical issue as they can cause `st_make_valid()` to fail, split into multi-geometries, or generate anomalous dimensions. We export this intermediate layer so we can revert back to it in case of issues.

```{r}
#| eval: false
#| code-summary: "Clean slivers"

# Calculate areas in the new projection
gk_data <- map(gk_data,
               ~mutate(.x, area_m2_proj = unclass(st_area(.x))))

# Identify polygons with area 0
gk_data_0 <- map(gk_data,
                 ~filter(.x, area_m2_proj =< 0.05)) # Random checks with QGIS
names(gk_data_0) <- str_extract(gdb_files, "(?<=format/).{2}")

# Clean them all after random checkes with QGIS
gk_data <- map(gk_data,
             ~filter(.x, area_m2_proj > 0.05)) 

# Export gk_data without sliver
# gpkg_path_sliver <- paste0(str_remove(tempdir(), "Rtmp\\w*"), "GK_raw/GK_proj_sliverscleaned_layer.gpkg")
# 
# iwalk(gk_data, function(layer, name) {
#         st_write(obj = layer,
#                  dsn = gpkg_path_sliver,
#                  layer = name,   
#                  append = FALSE,    
#                  delete_layer = TRUE)})

# Read intermediate layer (if needed)
# layer_names <- st_layers(gpkg_path_sliver)$name
# gk_data <- map(layer_names, ~st_read(gpkg_path_proj, layer = .x, quiet = TRUE))
# names(gk_data) <- str_extract(gdb_files, "(?<=format/).{2}")


# Clean R environment
rm(list=setdiff(ls(), c("pull",
                        "gpkg_path",
                        "gpkg_path_proj",
                        "gpkg_path_sliver",
                        "gk_data")))

```


Then we fix the geometry errors. We run `st_make_valid` and if it fails, we revert to `st_buffer(geometry, dist = 0)` which sometimes fixes the issues. As we expect that some empty geometries and geometrical issues will remain, we don't overwrite `gk_data` so we can further investigate if necessary. We will clean the R environment after the geometrical cleaning.
```{r}
#| eval: false
#| code-summary: "Fix geometry errors"

# Fix geometry
gk_data_valid <- gk_data %>%
                 imap(function(layer, name) {
                      tryCatch({
                        st_make_valid(layer)
                      }, error = function(e) {
                        message("st_make_valid() failed for layer '", name, "': ", conditionMessage(e),
                                "\nFalling back to st_buffer(dist = 0)")
                        st_buffer(layer, dist = 0)
                      })
                    })

# Export intermediate layer (if needed)
# gpkg_path_valid <- paste0(str_remove(tempdir(), "Rtmp\\w*"), "GK_raw/GK_proj_sliverscln_stvalid_layer.gpkg")
# 
# iwalk(gk_data_valid, function(layer, name) {
#         st_write(obj = layer,
#                  dsn = gpkg_path_valid,
#                  layer = name,   
#                  append = FALSE,    
#                  delete_layer = TRUE)})


# Read intermediate layer (if needed)
# layer_names <- st_layers(gpkg_path_valid)$name
# gk_data_valid <- map(layer_names, ~st_read(gpkg_path_valid, layer = .x, quiet = TRUE))
# names(gk_data_valid) <- str_extract(gdb_files, "(?<=format/).{2}")

# Clean R environment
rm(list=setdiff(ls(), c("pull",
                        "gpkg_path",
                        "gpkg_path_proj",
                        "gpkg_path_sliver",
                        "gpkg_path_valid",
                        "gk_data_valid",
                        "gk_data_valid")))

```


Then we do some checks to see if the corrections were efficient.
```{r}
#| eval: false
#| code-summary: "Check pre-cleaning worked"

valid_check <- lapply(gk_data_valid,
                      function(i){which(st_is_valid(i) == FALSE)})

empty_check <- lapply(gk_data_valid,
                      function(i){which(st_is_empty(i) == TRUE)})

lapply(valid_check, function (i){length(i)})
lapply(empty_check, function (i){length(i)})


```

Running `st_make_valid()` did not solve all of the issues. We still have 15 polygons with invalid geometries, and one polygon in tile 18 has an empty geometry despite an area of 20,156m2 prior to the cleaning step. We will first investigate the geometrical issues that could not be fixed. For that we directly ask what the problem is with `st_make_valid()`.

```{r}
#| eval: false
#| code-summary: "Investigate polygons with remaining geometrical issues"

# Polygons with unsolved geometrical issues
idx <- which(sapply(valid_check, length) > 0)

geom_issues <- map(idx, ~ gk_data_valid[[.x]][valid_check[[.x]], ])


# Reasons for not being valid
reasons <- geom_issues %>%
              map(.,
                  ~st_is_valid(.x, reason = TRUE))
```

The reason for the remaining invalidity is the self-intersection. We re-ran the function `st_make_valid()` and it fixed all the remaining geometrical issues. It is not clear as to why they were not fixed the first time around. We then replace these fixed geometry in the dataset.

```{r}
#| eval: false
#| code-summary: "Fixing self-intersections"

# Fix
geom_issues_fixed <- map(geom_issues,
                        ~st_make_valid(.x))

# Check the fixing worked
geom_issues_check <- map(geom_issues_fixed,
                          ~st_is_valid(.x))

# Replace fixed polygons in the original dataset
gk_data_valid_fixed <- imap(gk_data_valid, function(x, nm) {
  
  if (!nm %in% names(geom_issues_fixed)){ # don't touch if not in geom_issues_fixed
    return(x)
    
    }else{  
  
      st_geometry(x)[valid_check[[nm]]] <- st_geometry(geom_issues_fixed[[nm]])

      return(x)
      
}
  }
)

```

Now we can take care of the empty geometry. There is only one polygon with an empty geometry. Given its area, it cannot be deleted as it is not a sliver. We tried to re-run `st_make_valid()` but with the argument `geos_method = "valid_linework"` and the geometry was corrected. By default, `st_make_valid()` sets the argument `geos_method = "valid_structure"`. `geos_method = "linework"` which is a more aggressive clean-up that goes back to lines of the polygons and creates new nodes. But to do so, we had to correct the geometry of the polygon from the dataset `gk_data` where only the slivers had been cleaned. This is because `gk_data_valid`, on which `st_make_valid()` had been run already, returned an empty geometry. We need a greometry to run `st_make_valid` a second tine. It worked. We then just replaced the geometry of the polygon in the dataset.

```{r}
#| eval: false
#| code-summary: "Investigate and correct polygon with empty geometry"

# Polygon with empty geometry
empty_18 <- gk_data_valid$`18`[empty_check$`18`, ]

# Read tile 18 of geopackage GK_proj_sliverscleaned_layer
gk_data <- st_read(gpkg_path_sliver, layer = "18", quiet = TRUE)
names(gk_data_valid) <- "18"

# Repair the polygon with geos_method = "valid_structure"
empty_18_fixed <- st_make_valid(gk_data[empty_check$`18`,],
                                geos_method = "valid_linework")

# Replace fixed polygons in the original dataset
st_geometry(gk_data_valid_fixed$`18`)[empty_check$`18`] <- st_geometry(empty_18_fixed)

```

Now we can re-calculate all the areas based on the new valid geometries.

```{r}
#| eval: false
#| code-summary: "Re-calculate areas"

# Re-calculate areas
gk_data_valid_fixed <- map(gk_data_valid_fixed,
                           function(x){
                             
                             mutate(x, area_m2_final= unclass(st_area(x))) %>%
                             select(., !c(area_m2_proj, areal_m2))
                             }
                           ) 

# Export intermediate layer (if needed)
# gk_path_valid_fixed <- paste0(str_remove(tempdir(), "Rtmp\\w*"), "GK_raw/GK_valid_final.gpkg")
# 
# iwalk(gk_path_valid_fixed, function(layer, name) {
#         st_write(obj = layer,
#                  dsn = gpkg_path_valid,
#                  layer = name,   
#                  append = FALSE,    
#                  delete_layer = TRUE)})


# Read intermediate layer (if needed)
# gk_path_valid_fixed <- paste0(str_remove(tempdir(), "Rtmp\\w*"), "GK_raw/GK_valid_final.gpkg")
# 
# layer_names <- st_layers(gk_path_valid_fixed)$name
# 
# gk_data_valid_fixed <- map(layer_names, ~st_read(gk_path_valid_fixed, layer = .x, quiet = TRUE))
# 
# names(gk_data_valid_fixed) <- str_extract(gdb_files, "(?<=format/).{2}")


# Clean R environment
rm(list=setdiff(ls(), c("pull",
                        "gpkg_path",
                        "gpkg_path_proj",
                        "gpkg_path_sliver",
                        "gpkg_path_valid",
                        "gk_path_valid_fixed",
                        "gk_data_valid_fixed")))


```




# BALINT ADDITIONAL CLEANING STEPS. 
I didn't do any union before as I thought you might need all the small polygons? I also left it as a list where each item is a tile. 


Now we merge all the tiles together, check for validity (making sure that all the tiles fit together nicely) and dissolve the ecosystem types into single polygons for further analyses. This should enable us to save some computing time. 
```{r}
#| eval: false
#| code-summary: "Dissolve ecosystem types into one single polygon"

# Merge ET level one per tiles 
gk_data_final <- gk_data_valid_fixed %>% 
                    map(., function(x){
                      x %>%
                        group_by(okosystemtype_1) %>%
                       summarise(geom = st_union(geom),
                                 area_m2_total = sum(area_m2_final), 
                                 .groups = "drop")
                    }) 


# Save intermediate product (if needed)
# gk_path_data_join <- paste0(str_remove(tempdir(), "Rtmp\\w*"), "GK_raw/GK_ETlevel1_union_tiles_21JULY2026.gpkg")
# 
# iwalk(gk_path_data_join, function(layer, name) {
#         st_write(obj = layer,
#                  dsn = gk_path_data,
#                  layer = name,   
#                  append = FALSE,    
#                  delete_layer = TRUE)})

# Read intermediate layer (if needed)
# gk_path_data_join <- paste0(str_remove(tempdir(), "Rtmp\\w*"), "GK_raw/GK_ETlevel1_union_tiles_21JULY2026.gpkg")
# 
# layer_names <- st_layers(gk_path_data_join)$name
# 
# gk_data_final <- map(layer_names, ~st_read(gk_path_data_join, layer = .x, quiet = TRUE))
# 
# names(gk_data_final) <- str_extract(gdb_files, "(?<=format/).{2}")

# Merge all tiles together and union ET level 1 at national level
gk_data_final <- gk_data_final %>% # overwrite data to spare memory
                   compact() %>%  
                   bind_rows() %>%
                   group_by(okosystemtype_1) %>%
                   summarise(geom = st_union(geom),
                             area_m2_total = sum(area_m2_final), 
                             area_m2_st = unclass(st_area(.)),
                             .groups = "drop")

# Check validity
final_validity <- which(st_is_valid(gk_data_final) == FALSE)

# Export
gk_path_data_final <- paste0(str_remove(tempdir(), "Rtmp\\w*"), "GK_raw/GK_final_19JULY2026.gpkg")

st_write(gk_data_final, gk_path_data_final, layer = "ET level 1")


st_write(filter(gk_data_final, okosystemtype_1 == ""), 
         gk_data_final, 
         layer = "my_layer_name")

st_write(my_sf_object, gk_data_final, layer = "my_layer_name", append = TRUE)

# Clean R Environment
rm(list=setdiff(ls(), c("pull",
                        "gpkg_path",
                        "gpkg_path_proj",
                        "gpkg_path_sliver",
                        "gpkg_path_valid",
                        "gk_path_valid_fixed",
                        "gk_data_final")))


```


## 6. Export file
We export the data as a geopackage where each layer is an ecosystem type.
```{r export}
#| eval: false



```

## 7. References
<!--# You can add references manually or use a citation manager and add intext citations as with crossreferencing and hyperlinks. See https://quarto.org/docs/authoring/footnotes-and-citations.html -->


::: {#refs}
:::

## 8. Author contributions

<!-- Fill in CReDIT and MeRIT statments to clarify the contributions from individual authors. For MeRIT, see https://www.nature.com/articles/s41467-023-37039-1. For CRedit, see https://credit.niso.org/. You can start with the template structur below, and just delete the rows you dont use.  -->

**CRediT statement**

* Conceptualization: SC
* Methodology: SC & BC
* Data curation & formatting: SC
* Writing – original draft: SC
* Writing – review & editing: BC
* Document structure: SC & AK


**MeRIT attribution**

The structure of the document has been taken from ecRxiv's template for condition indicators, created by Anders Kolstad (AK). SC adapted it to the purpose of the project.

**Study design**
The study was designed by SC and BC.

**Data processing**
Raw data were downloaded by SC. They were curated and cleaned by SC and BC. Code was written by both SC and BC.


::: {.callout-tip collapse="true"}
## Session Info
<!-- You can leave this last part as it is, unless you have a good reason to alter it. -->

This information is mainly aimed at developers that need to recreate the R environment that produced the html.
Important note: this information is about the R session that was running when the html was rendered, and if some of the code chunks were not evaluated when the html was rendered, they may have been executed in a different environment.  

```{r sessionInfo}
#| echo: false
sessionInfo()
```

:::
