---
title: "Prepare Copernicus High Resolution Layer Tree Cover Density data"
folderName: code # Leave to be filled out by ecRxiv during typesetting (name of the folder under 'indicators/'). 
# If it's a single indicator being reported, the folder name should be the same as the indicatorID. 
# If there are multiple ID's then the folder name could look something like XX_YYY_ZZZ-ÅÅÅ.
country: Norway
continent: Europe
yearAdded: 2026 # the year when the indicator was first submitted to ecRxiv
yearLastUpdate: 2026
status: complete # options are complete, incomplete, or deprecated
Version: "0.1.0" # See wiki page: https://github.com/NINAnor/ecRxiv/wiki#versioning
VersionComment: 'First version. The code has been adapted from https://github.com/NINAnor/ecRxiv/tree/main/indicators/NO_TCCD_001_104. Template adapated from Anders Kolstad ecRxiv repository: https://github.com/NINAnor/ecRxiv.'
url: https://github.com/NINAnor/egd_data_preparation
# To be filled out by ecRxiv during the typesetting, when they will replace XX_YYYY_ZZ with the name of the root folder for the current indicator(s). It should be the same as the indicatorID, but if there are more indicator IDs presented in the same folder, the folder name might be different, e.g. XX_YYYY_AAA-BBB 
format: 
  html:
    embed-resources: true
    code-fold: true
    toc: true
    toc-title: Contents
    toc-depth: 3
    smooth-scroll: true
execute: 
  cache: true
author:
  - name: Sylvie Clappe             # Enter name
    email: sylvie.clappe@nina.no  # Enter email
    affiliations:
      - id: myID
        name: Norwegian Institute for Nature Research # Enter affiliations
  - name: Bálint Czúcz              
    email: balint.czucz@nina.no             #  Enter subsequent authors like this, or remove if not relevant
    affiliations:
      - ref: myID               # To reuse affiliations reference the id like this
AuthorList: Clappe, S., Czúcz, B.
date: August 11, 2026
callout-icon: false
lightbox: true
css: ../../../style.css
code-links:
      - text: Add a review
        icon: github
        href: https://github.com/NINAnor/ecRxiv
hide: TRUE # Change to FALSE
---

<!--# This is a template for how to document the indicator analyses. Make sure also to not change the order, or modify, the headers, unless you really need to. This is because it easier to read if all the indicators are presented using the same layout. If there is one header where you don't have anything to write, just leave the header as is, and don't write anything below it. If you are providing code, be careful to annotate and comment on every step in the analysis. Before starting it is recommended to fill in as much as you can in the metadata file. This file will populate the initial table in your output.-->

<!--# Load all you dependencies here -->

```{r setup}
#| include: false
#| cache: false
library(knitr)
library(tidyverse)
library(kableExtra)
library(here)
library(yaml)
library(tibble)
library(conflicted)
library(tidyverse)
library(dplyr)
library(terra)
library(carrier)
library(mirai)
library(hdar)
library(jsonlite)
library(DBI)
library(duckspatial)
library(sf)
library(httr2)

# weird workarounds for cache issues
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::lag)
conflicts_prefer(dplyr::select)
pull<-dplyr::pull

knitr::opts_chunk$set(echo = TRUE)
```

<!--# The following parts are autogenerated. Do not edit. -->


```{r source}
#| echo: false
#| cache: false

source(here::here("_common.R"))
```


```{r reviewers}
#| echo: false
#| results: asis

if (!is.null(yaml_data$codeReviewers)) {
  reviewers <- purrr::map_chr(yaml_data$codeReviewers, "name")
  
  htmltools::HTML(paste0(
    '<div id="title-block-header" class="quarto-title-block default">
       <div class="quarto-title-meta">
         <div>
           <div class="quarto-title-meta-heading">Reviewed by</div>
           <div class="quarto-title-meta-contents">',
    paste(reviewers, collapse = "; "),
    '</div>
         </div>
       </div>
     </div>',
    sep = ""
  ))
}
```

> **Recommended citation**: `r paste(auth, " ", year, ". ", "v. ", version, ": ", url, sep="")`



::: {.callout-tip collapse="true"}
## Metadata
```{r tbl-meta2}
#| tbl-cap: 'Indicator metadata'
#| echo: false
#| warning: false
meta |>
  dplyr::filter(Variable %in% c(
    "Indicator ID",
    "Indicator Name",
    "verbatimName",
    "Continent",
    "Country",
    "Year added", 
    "Last update",
    "Version",
    "Version comment")) |>   
  kable()

```
:::

::: {.callout-tip collapse="true"}

## Log

<!--# Update this logg with short messages for each update -->
- 11 August 2026 - First draft version
:::


<hr />

<!--# Document you work below.  -->

## 1. Summary
This code shows how to download and pre-process Copernicus High Resolution Layer (HRL) Tree Cover Density (TCD) data to compile the TCCD indicators from ecRxiv: https://github.com/NINAnor/ecRxiv/tree/main/indicators/NO_TCCD_001_104.


## 2. Data characteristics
These data are produced by Copernicus: https://land.copernicus.eu/en/products/high-resolution-layer-forests-and-tree-cover.

### 2.1 Spatial and temporal resolution and extent
We downloaded the data for the whole on Norway (excluding Svalbård), for the years 2018, 2019, 2020, 2021. Spatial resolution is 10m x 10m.

### 2.2 Sampling design
The HRL TCD data are generated using state-of-the-art automated image processing 
methods that are applied to systematically acquired, high-resolution satellite image time series 
from the Copernicus space segment (i.e. Sentinel-2). Layers have been harmonised across time, allowing for time series comparisons. See [Algorithm Teoretical Basis Document](https://library.land.copernicus.eu/products/High_Resolution_Layer_Vegetated_Land_Cover_Characteristics_ATBD_v2.html) and [Product user manual](https://land.copernicus.eu/en/technical-library/product-user-manual-tree-cover-and-forests-2018-present/@@download/file).


### 2.3 Original units
The original unit of Copernicus HRL TCD is percentages (the share of tree cover areas in the pixel).


### 2.4 Instructions for citing, using and accessing data
Openly available[here](https://land.copernicus.eu/en/products/high-resolution-layer-forests-and-tree-cover/tree-cover-density-2018-present-raster-10-m-europe-yearly#download)


### 2.5 Additional comments about the dataset
Other metadata can be found in the [Product user manual](https://land.copernicus.eu/en/technical-library/product-user-manual-tree-cover-and-forests-2018-present/@@download/file)


## 3. Uncertainties
### 3.1 General uncertainties
Uncertainties have several sources all taking roots in the modelling used to process the Sentinel-2 raw data. They can come from the pre-processing of the raw data necessary to correct the images (e.g., radiometrical/sensor calibration; geophysical corrections; cloud cover corrections;creation of the mosaic). But they also arise when the pre-processed data are classified through machine learning to produce the final layer of imperviousness. Classification is never definitive and produces probabilities for a pixel to belong to a specific class. 

### 3.2 Reported uncertainties
Copernicus reported "Confidence layers" for each of the TCD datasets. In their technical guide, they explain that, they report the "Total Uncertainty" which is defined as the sum of data and knowledge uncertainty. "data uncertainty" is a measure for the data complexity due to noise or overlapping classes. While "knowledge uncertainty" is obtained through the CatBoost algorithm by using virtual ensembles. Basically, CatBoost runs on several data ensembles and compiles the mean value. THe "Knowledge uncertainty" is then the variance of the al the means comiled.

The confidence values are pixel-based and range from 0 to 100. In practice only values 50-100 are used as values below 50% would indicate a higher confidence in the opposite class. The recommended interpretation state that pixels with values ranging 75-100 are of high confidence, 50-75 of moderate confidence, and below 50 of low confidence.

The confidence layers are available for all years.


## 4. Data download

### 4.1 Copernicus TCD
Tree canopy cover density layers can be downloaded from Copernicus for the years 2018 to 2023. This can be done using an API query through the [WEkEO website](https://help.wekeo.eu/en/articles/6416936-how-to-download-wekeo-data#h_4d7de19c7c) and the `hda` R package. A detailed tutorial can be found [here](https://help.wekeo.eu/en/articles/7035318-how-to-use-the-hdar-package-for-accessing-the-wekeo-hda-api-in-r).Note that the request through API requires the username and password from the WEkEO website so you nee dto create an account. The API query uses a bounding box but it is quite large, resulting in heavy rasters. That is why a subsequent step of cleaning has been done to mask the cells that were not in Norway, hence decreasing the size of the rasters. These processed rasters are the ones we made available on the internal NINA repository.

*!Attention: this code, writes the rasters on the disk of the users.*

First we need to set up our WekEO credentials.

```{r}
#| eval: false
#| code-summary: "Set your WEkEO credentials"

## Download data
# Set credentials and save them in ~/.hdarc for future use
username <- "your_username"
password <-"your_password"
client <- Client$new(username, password, save_credentials = TRUE)

# Initialise Client
client <- Client$new()

# Check authentification
client$get_token() 

# Review and accept Terms & Condition before downloading the data
client$show_terms()
client$terms_and_conditions(term_id = 'all')

```

Now we can create the API query. It can be automatically created for you using the [Expert Data Viewer](https://data.wekeo.copernicus.eu/data?view=viewer&objects=W10%3D&layers=eyJjMCI6eyJpZCI6ImMwIiwib3BlbkRyYXdlck5hbWUiOiJzdWJzZXR0ZXIiLCJyZXBsYWNlbWVudENvbG9yTWFwSWQiOm51bGwsInZhbHVlTWluIjpudWxsLCJ2YWx1ZU1heCI6bnVsbCwibGF5ZXJJZCI6IkVPOkVFQTpEQVQ6SFJMOlRDRi9fX0RFRkFVTFRfXy90cmVlQ292ZXIiLCJ6SW5kZXgiOjIwLCJpc0luaXRpYWwiOnRydWUsImlzVmlld2FibGUiOnRydWUsInRpbWUiOjE2NzI1MzEyMDAwMDAsImlzUmVwbGFjaW5nIjpmYWxzZX19). There is an additional trick for HRL TCD though. The API format of these layers have changed and the Expert Data Viewer does not reflect this. It is still useful to take the basic API from the Expert Data Viewer and then modify it with the instructions [here](https://help.wekeo.eu/en/articles/15612836-action-required-changes-to-data-access-for-clms-hrl-datasets).

```{r}
#| eval: false
#| code-summary: "Create the API query"

# Create API queries
# Note: use the WEkEO Data viewer that will create an automatic API
# Note: Use "SetAOI" to set your bounding box
yrs <- list(c("2022-01-01T00:00:00.000Z", 
              "2022-12-31T00:00:00.000Z"),
            c("2023-01-01T00:00:00.000Z", 
              "2023-12-31T00:00:00.000Z"))

api_list <- map(seq_along(yrs), function(i){
  paste0('{
  "dataset_id": "EO:JRC:DAT:CLMS",
  "bbox": [
    4.035020510000001,
    57.61991392,
    32.10650033,
    71.32500159
  ],
  "startdate": "',
  yrs[[i]][1],
  '", 
  "enddate": "', 
  yrs[[i]][2],
  '",
  "datasetIdentifier": "clms_vlcc_tree-cover-density_europe_10m_yearly_v1",
  "itemsPerPage": 200,
  "startIndex": 0
}')}
) %>% 
  unlist()

```

We use the API query to search and download the data sets of relevance.
```{r}
#| eval: false
#| code-summary: "Search and dowload the data corresponding to the API query"

# Search the datasets
matches <- map(seq_along(api_list), ~ client$search(api_list[.x]))
names(matches) <- c("2022", "2023")

# ID of all the datasets to download
data_id <- map(matches, function(yr) {
  map_dfr(yr$results, ~ tibble(id = .x$id))
})

# Set up the output directory
output_directory <- paste0(str_remove(tempdir(), "Rtmp\\w*"), "TCD_Copernicus")

# Download - co-written by Sylvie Clappe and Jennifer Hansen
walk(matches, function(yr) {
  
 walk(seq_along(yr$results)[51:172], function(i) {
  
  product_id <- yr$results[[i]]$id
  
  message(
    sprintf(
      "Downloading %s %d/%d: %s",
      str_extract(product_id, "(?<=TCD_S).{4}"),
      i,
      yr$total_count,
      product_id
    )
  )
  
  tryCatch(
    {
      yr$download(output_dir = output_directory,
                  selected_indexes = i,
                  prompt = FALSE) },
    error = function(e) {
      message("Failed: ", product_id)
      message(conditionMessage(e))
    }
  )
  
  Sys.sleep(1) # to avoid hitting API quota
 })
})

# Clean R environment
rm(list = setdiff(ls(), 
                  c("pull",
                    "client")))


```

### 4.2 Copernicus IMD Confidence layers
Now we do the same for the confidence layers.

```{r}
#| eval: false
#| code-summary: "Create the API query"

# Create API queries
# Note: use the WEkEO Data viewer that will create an automatic API
# Note: Use "SetAOI" to set your bounding box
yrs <- list(c("2018-01-01T00:00:00.000Z", 
              "2018-12-31T00:00:00.000Z"),
            c("2019-01-01T00:00:00.000Z", 
              "2019-12-31T00:00:00.000Z"),
            c("2020-01-01T00:00:00.000Z", 
              "2020-12-31T00:00:00.000Z"),
            c("2021-01-01T00:00:00.000Z", 
              "2021-12-31T00:00:00.000Z"),
            c("2022-01-01T00:00:00.000Z", 
              "2022-12-31T00:00:00.000Z"),
            c("2023-01-01T00:00:00.000Z", 
              "2023-12-31T00:00:00.000Z"))

api_list <- map(seq_along(yrs), function(i){
  paste0('{
  "dataset_id": "EO:JRC:DAT:CLMS",
  "bbox": [
    4.035020510000001,
    57.61991392,
    32.10650033,
    71.32500159
  ],
  "startdate": "',
  yrs[[i]][1],
  '", 
  "enddate": "', 
  yrs[[i]][2],
  '",
  "datasetIdentifier": "clms_vlcc_tree-cover-density-confidence-layer_europe_10m_yearly_v1",
  "itemsPerPage": 200,
  "startIndex": 0
}')}
) %>% 
  unlist()

```

```{r}
#| eval: false
#| code-summary: "Search and dowload the data corresponding to the API query"

# Search the datasets
matches <- map(seq_along(api_list), ~ client$search(api_list[.x]))
names(matches) <- as.character(seq(2018, 2023, 1))

# ID of all the datasets to download
data_id <- map(matches, function(yr) {
  map_dfr(yr$results, ~ tibble(id = .x$id))
})

# Set up the output directory
output_directory <- paste0(str_remove(tempdir(), "Rtmp\\w*"), "TCD_Copernicus_confidence")

# Download - co-written by Sylvie Clappe and Jennifer Hansen
walk(matches, function(yr) {
  
 walk(seq_along(yr$results), function(i) {
  
  product_id <- yr$results[[i]]$id
  
  message(
    sprintf(
      "Downloading %s %d/%d: %s",
      str_extract(product_id, "(?<=TCDCL_S).{4}"),
      i,
      yr$total_count,
      product_id
    )
  )
  
  tryCatch(
    {
      yr$download(output_dir = output_directory,
                  selected_indexes = i,
                  prompt = FALSE) },
    error = function(e) {
      message("Failed: ", product_id)
      message(conditionMessage(e))
    }
  )
  
  Sys.sleep(1) # to avoid hitting API quota
 })
})

# Clean R environment
rm(list = setdiff(ls(), 
                  c("pull")))


```

## 5. Data cleaning and formatting
Once the rasters (.tf) have been extracted and written on the disk, they can be read to create a virtual raster that will be used for the subsequent analyses later on.

### 5.1 Copernicus TCD
All the data have been downloaded as `.zip`. We are now going to unzip them. While we do it, we perform additional cleaning and formatting steps. Norway bounding box for which we downloaded the data is quite wide and include a lot of rasters that are in Sweden and Finland, while not overlapping Norway. We apply a mask for Norway to select the relevant rasters. For this we use the administrative boundaries from Geonorge. We also re-classify the raster value of 255 as NaN.

First, we create a mask of Norway.
```{r}
#| eval: false
#| code-summary: "Load the administrative boundaries"

# Call function
source(here::here("src", "functions", "api_geonorge.R"))

# Open duckdb connection
conn_tcd <- ddbs_create_conn(dbidr = "tempdir", threads = 5, memory_limit_gb = 10)

# Download data in tempdir
no_data_path <- api_geonorge("https://nedlasting.geonorge.no/api/capabilities/041f1e6e-bdbc-4091-b48f-8a5990f3cc5b", 
                            "john.doe@email.no")

st_layers(no_data_path)

# Read, transform, union and write 
dbExecute(conn_tcd, 
            sprintf("
              CREATE OR REPLACE TABLE '%s' AS
              SELECT ST_Transform(
             ST_Union_Agg(ST_MakeValid(SHAPE)),
             'EPSG:25833',
             'EPSG:3035',
             always_xy := true) AS geometry
              FROM ST_Read('%s', layer = '%s')
            ", 
      "norway_bd",
      no_data_path,
      "kommune"))

# Extract dataset
norway_mask <- as_duckspatial_df("norway_bd", conn = conn_tcd) %>%
                ddbs_collect() %>%
                st_set_crs(3035) # duckdb doesn't track the change of the CRS in the database so needs to be done manually

# Verify tables are in my duckdb conn_pm database
ddbs_stop_conn(conn_tcd)

# Clean the R environment
rm(list = setdiff(ls(), 
                  c("pull",
                    "output_directory",
                    "norway_mask")))

```

Now that the mask is ready, we can unzip the files and clean the rasters.

```{r}
#| eval: false
#| code-summary: "Read the .tif from a zip file"

# Call functions
source(here::here("src", "functions", "list_zip_files.R"))
source(here::here("src", "functions", "extract_zip_files.R"))


# List datasets' paths to read and their names
files_tcd_read <- files_list_one_fn(output_directory)

# List datasets path (here we remove the zip marker and the name of the file)
files_tcd_write <- str_remove(files_tcd_read[[1]], "CLMS.*") %>%
                      str_remove("/vsizip/\\{")

# Run the function to read and write the rasters from the zip files directly
daemons(10)
map(seq_along(files_tcd_read[[1]]),
    in_parallel( ~process_raster_fn(path_read = files_tcd_read[[1]][.x],
                                    path_write = files_tcd_write[.x],
                                    files_names = files_tcd_read[[2]][.x],
                                    geo_mask = norway_mask),
                 process_raster_fn = process_raster_fn,
                 files_tcd_read = files_tcd_read,
                 files_tcd_write = files_tcd_write,
                 norway_mask = norway_mask
    )
)

daemons(0)

# Clean the R environment
rm(list = setdiff(ls(), 
                  c("pull",
                    "output_directory")))

```

We now create a virtual raster that we will read and export as a raster for future analyses. These rasters are not heavy (<2GB each).

```{r}
#| eval: false
#| code-summary: "Create Copernicus IMD virtual rasters and export them as rasters"

## Code co-created by Sylvie Clappe and Jennifer Hansen

## Create the IMD virtual rasters
file_tcd_list <- dir(output_directory, recursive = FALSE, pattern = ".tif$", full.names = TRUE)
yrs_tcd <- str_extract(file_tcd_list, "(?<=TCD_S).{4}") %>%
              unique()

file_tcd <- map_vec(seq_along(yrs_tcd), function(i){paste0(output_directory, "/TCD_", yrs_tcd[i], ".vrt")})

walk(seq_along(file_tcd), function(x){
  
  # Select relevant .tif
    file_tcd_list[grepl(yrs_tcd[x], file_tcd_list, ignore.case = TRUE)] %>%
    
  # Create virtual raster
    vrt(filename= file_tcd[x]) } )
               
        
## Read .vrt files
rast_list <- map(seq_along(file_tcd),
                 ~rast(file_tcd[.x]))

## Add levels 
levels_original <- levels(rast(file_tcd_list[1]))
levels_original[[1]]$class_name[levels_original[[1]]$class_name == "outside area"] <- "outside area reclassified as NAs"

rast_list <- map(rast_list,
                 ~{levels(.x) <- levels_original
                 return(.x)})

## Export new rasters
walk(seq_along(file_tcd),
     ~writeRaster(rast_list[[.x]], filename = paste0(output_directory, "/TCD_", yrs_tcd[.x], ".tif")))

```


### 5.2 Copernicus IMD Confidence layers
```{r}
#| eval: false
#| code-summary: "Read the .tif from a zip file"

# Call functions
source(here::here("src", "functions", "list_zip_files.R"))
source(here::here("src", "functions", "extract_zip_files.R"))

# Set up the output directory
output_directory <- paste0(str_remove(tempdir(), "Rtmp\\w*"), "TCD_Copernicus_confidence")

# List datasets' paths to read and their names
files_tcd_read <- files_list_one_fn(output_directory)

# List datasets path (here we remove the zip marker and the name of the file)
files_tcd_write <- str_remove(files_tcd_read[[1]], "CLMS.*") %>%
                      str_remove("/vsizip/\\{")

# Run the function to read and write the rasters from the zip files directly
daemons(10)
map(seq_along(files_tcd_read[[1]]),
    in_parallel( ~process_raster_fn(path_read = files_tcd_read[[1]][.x],
                                    path_write = files_tcd_write[.x],
                                    files_names = files_tcd_read[[2]][.x],
                                    geo_mask = norway_mask),
                 process_raster_fn = process_raster_fn,
                 files_tcd_read = files_tcd_read,
                 files_tcd_write = files_tcd_write,
                 norway_mask = norway_mask
    )
)

daemons(0)

# Clean the R environment
rm(list = setdiff(ls(), 
                  c("pull",
                    "output_directory")))

```

```{r}
#| eval: false
#| code-summary: "Create Copernicus IMD virtual rasters and export them as rasters"

## Code co-created by Sylvie Clappe and Jennifer Hansen

## Create the IMD virtual rasters
file_tcd_list <- dir(output_directory, recursive = FALSE, pattern = ".tif$", full.names = TRUE)
yrs_tcd <- str_extract(file_tcd_list, "(?<=TCDCL_S).{4}") %>%
              unique()

file_tcd <- map_vec(seq_along(yrs_tcd), function(i){paste0(output_directory, "/TCDCL_", yrs_tcd[i], ".vrt")})

walk(seq_along(file_tcd), function(x){
  
  # Select relevant .tif
    file_tcd_list[grepl(yrs_tcd[x], file_tcd_list, ignore.case = TRUE)] %>%
    
  # Create virtual raster
    vrt(filename= file_tcd[x]) } )
               
        
## Read .vrt files
rast_list <- map(seq_along(file_tcd),
                 ~rast(file_tcd[.x]))

## Add levels 
levels_original <- levels(rast(file_tcd_list[1]))
levels_original[[1]]$class_name[levels_original[[1]]$class_name == "outside area"] <- "outside area reclassified as NAs"

rast_list <- map(rast_list,
                 ~{levels(.x) <- levels_original
                 return(.x)})

## Export new rasters
walk(seq_along(file_tcd),
     ~writeRaster(rast_list[[.x]], filename = paste0(output_directory, "/TCDCL_", yrs_tcd[.x], ".tif")))

```

## 6. Export file
Virtual raster already exported in the previous code.

## 7. Author contributions


**CRediT statement**

* Conceptualization: SC
* Methodology: SC & BC
* Data curation & formatting: SC
* Writing – original draft: SC
* Writing – review & editing: BC
* Document structure: SC & AK


**MeRIT attribution**

The structure of the document has been taken from ecRxiv's template for condition indicators, created by Anders Kolstad (AK). SC adapted it to the purpose of the project.

**Study design**
The study was designed by SC and BC.

**Data processing**
Raw data were downloaded by SC. They were curated and cleaned by SC and BC. Code was written by both SC and BC.


::: {.callout-tip collapse="true"}
## Session Info
<!-- You can leave this last part as it is, unless you have a good reason to alter it. -->

This information is mainly aimed at developers that need to recreate the R environment that produced the html.
Important note: this information is about the R session that was running when the html was rendered, and if some of the code chunks were not evaluated when the html was rendered, they may have been executed in a different environment.  

```{r sessionInfo}
#| echo: false
sessionInfo()
```

:::
