---
title: "Prepare Particulate Matter data"
folderName: code # Leave to be filled out by ecRxiv during typesetting (name of the folder under 'indicators/'). 
# If it's a single indicator being reported, the folder name should be the same as the indicatorID. 
# If there are multiple ID's then the folder name could look something like XX_YYY_ZZZ-ÅÅÅ.
country: Norway
continent: Europe
yearAdded: 2026 # the year when the indicator was first submitted to ecRxiv
yearLastUpdate: 2026
status: complete # options are complete, incomplete, or deprecated
Version: "0.1.0" # See wiki page: https://github.com/NINAnor/ecRxiv/wiki#versioning
VersionComment: 'First version. The code has been adapted from https://github.com/NINAnor/ecRxiv/tree/main/indicators/NO_URAQ_001-005. Template adapated from Anders Kolstad ecRxiv repository: https://github.com/NINAnor/ecRxiv.'
url: https://github.com/NINAnor/egd_data_preparation
# To be filled out by ecRxiv during the typesetting, when they will replace XX_YYYY_ZZ with the name of the root folder for the current indicator(s). It should be the same as the indicatorID, but if there are more indicator IDs presented in the same folder, the folder name might be different, e.g. XX_YYYY_AAA-BBB 
format: 
  html:
    embed-resources: true
    code-fold: true
    toc: true
    toc-title: Contents
    toc-depth: 3
    smooth-scroll: true
execute: 
  cache: true
author:
  - name: Sylvie Clappe             # Enter name
    email: sylvie.clappe@nina.no  # Enter email
    affiliations:
      - id: myID
        name: Norwegian Institute for Nature Research # Enter affiliations
  - name: Bálint Czúcz              
    email: balint.czucz@nina.no             #  Enter subsequent authors like this, or remove if not relevant
    affiliations:
      - ref: myID               # To reuse affiliations reference the id like this
AuthorList: Clappe, S., Czúcz, B.
date: July 20, 2026
callout-icon: false
lightbox: true
css: ../../../style.css
code-links:
      - text: Add a review
        icon: github
        href: https://github.com/NINAnor/ecRxiv
hide: TRUE # Change to FALSE
---

<!--# This is a template for how to document the indicator analyses. Make sure also to not change the order, or modify, the headers, unless you really need to. This is because it easier to read if all the indicators are presented using the same layout. If there is one header where you don't have anything to write, just leave the header as is, and don't write anything below it. If you are providing code, be careful to annotate and comment on every step in the analysis. Before starting it is recommended to fill in as much as you can in the metadata file. This file will populate the initial table in your output.-->

<!--# Load all you dependencies here -->

```{r setup}
#| include: false
#| cache: false
library(knitr)
library(tidyverse)
library(kableExtra)
library(here)
library(yaml)
library(tibble)
library(conflicted)
library(tidyverse)
library(dplyr)
library(terra)
library(carrier)
library(mirai)
library(giscoR)

# weird workarounds for cache issues
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::lag)
conflicts_prefer(dplyr::select)
pull<-dplyr::pull

knitr::opts_chunk$set(echo = TRUE)
```

<!--# The following parts are autogenerated. Do not edit. -->


```{r source}
#| echo: false
#| cache: false

source(here::here("_common.R"))
```


```{r reviewers}
#| echo: false
#| results: asis

if (!is.null(yaml_data$codeReviewers)) {
  reviewers <- purrr::map_chr(yaml_data$codeReviewers, "name")
  
  htmltools::HTML(paste0(
    '<div id="title-block-header" class="quarto-title-block default">
       <div class="quarto-title-meta">
         <div>
           <div class="quarto-title-meta-heading">Reviewed by</div>
           <div class="quarto-title-meta-contents">',
    paste(reviewers, collapse = "; "),
    '</div>
         </div>
       </div>
     </div>',
    sep = ""
  ))
}
```

> **Recommended citation**: `r paste(auth, " ", year, ". ", "v. ", version, ": ", url, sep="")`



::: {.callout-tip collapse="true"}
## Metadata
```{r tbl-meta2}
#| tbl-cap: 'Indicator metadata'
#| echo: false
#| warning: false
meta |>
  dplyr::filter(Variable %in% c(
    "Indicator ID",
    "Indicator Name",
    "verbatimName",
    "Continent",
    "Country",
    "Year added", 
    "Last update",
    "Version",
    "Version comment")) |>   
  kable()

```
:::

::: {.callout-tip collapse="true"}

## Log

<!--# Update this logg with short messages for each update -->
- 13 July 2026 - First draft version
:::


<hr />

<!--# Document you work below.  -->

## 1. Summary
This code shows how to download and pre-process PM2.5 data from the Norwegian Meteorological Institute. PM2.5 data will serve to compile compile the URAQ indicators from ecRxiv: https://github.com/NINAnor/ecRxiv/tree/main/indicators/NO_URAQ_001-005.


## 2. Data characteristics

These data are produced by the Norwegian Meteorological Institute whose data catalogue is found [here](https://thredds.met.no/thredds/catalog/data/fou-kl/uEMEP/for_NTNU/catalog.html).

### 2.1 Spatial and temporal resolution and extent

We downloaded the data for the whole on Norway (excluding Svalbård), yearly from 2016 to 2024. These are raster data with a spatial resolution of 100m x 100m. These data represent the annual mean.

### 2.2 Sampling design
The annual PM2.5 dataset a model-based gridded product generated  using a nested air-quality modelling framework using EMEP and uEMEP. EMEP is a chemistry and transport model, while uEMEP is a dispersion model used at local scale. Input data include emissions, meteorological data from a weather model, and geographic and land-surface data. Monitoring stations were only used in validation of the model.

The PM2.5 concentrations are calculated hourly on a national 100 m × 100 m grid, and the annual PM2.5 raster is then produced by averaging the hourly concentrations over the entire year for each grid cell. See metadata [here](https://www.met.no/prosjekter/luftkvalitet/dokumentasjon-av-luftkvalitetsmodellen).



### 2.3 Original units
The original unit is µg/m³ (micrograms of particulate matter smaller than 2.5 µm per cubic metre of air)


### 2.4 Instructions for citing, using and accessing data
Openly available [here](https://thredds.met.no/thredds/catalog/data/fou-kl/uEMEP/for_NTNU/catalog.html)


### 2.5 Additional comments about the dataset
Other metadata can be found in the [here](https://www.met.no/prosjekter/luftkvalitet/dokumentasjon-av-luftkvalitetsmodellen)


## 3. Uncertainties
### 3.1 General uncertainties
Uncertainties are not quantified in the data, but they are discussed qualitatively identifying six main sources:

- errors in emissions inventories
- modelling assumptions
- model structure
- uncertainties of the datasets used for modelling inputs

### 3.2 Reported uncertainties
Uncertainties are not quantified in the data, but they are discussed qualitatively. 


## 4. Data download
Data can be downloaded from the Norwegian Meteorological Institute catalogue [here](https://thredds.met.no/thredds/catalog/data/fou-kl/uEMEP/for_NTNU/catalog.html). We download them manually for years 2016 to 2024. from met.no catalogue. These data could not be loaded in the R environment directly through an API, they could only be downloaded (at the time the code was developed). We stored them on NINA's internal shared drive. To use our code more easily, we invite the users to download the data and modify the data path we provide below. The data are not provided as rasters but not as .nc, not .tif. THis is because .nc is more standard for climate data. 

```{r}
#| eval: false
#| code-summary: "PM2.5 data path"

pm25_path <- "/data/P-Prosjekter2/412413_2023_no_egd/git_data/PM2.5/National"

```

We have nine rasters to load, we first create a list of files before reading them and storing them in a list object. *Note: the CRS of the layers could not be identified precisely by R "unnamed". After contacting the data provider, we had the confirmation that the CRS was EPSG:32633*

```{r}
#| eval: false
#| code-summary: "Read and stor PM2.5 data"
# List the gdb files
pm25_files <- list.files(pm25_path, pattern = ".nc$", 
                         full.names = TRUE)

# Read data with gdb files
pm25_data <- pm25_files %>% 
                 map(\(.) rast(.)) %>%
                 map(., ~ {
                   crs(.x) <- "EPSG:32633"
                   .x
                   })

# Name the layers
names(pm25_data) <- str_extract(pm25_files, "(?<=annualmean_).{4}")

```


## 5. Data cleaning and formatting
Once the rasters (.tf) have been loaded in the environment, we can further format them. We need to mask them with the terrestrial administrative boundaries. As is, the rasters include part of the marine ecosystems. The URAQ condition indicators are restricted to terrestrial ecosystem only (ET01 - Settlements and other artificial areas). We assume that the administrative boundaries exclude the marine ecosystems. 

```{r}
#| eval: false
#| code-summary: "Mask PM2.5 data"

# Load administrative boundaries
# Create a temporary directory to store the cache
temp_dir <- tempdir()

# Download the dataset
no_mask <- gisco_get_lau(year = "2024",
                        epsg = "3035",
                        update_cache = TRUE,
                        cache_dir = temp_dir,
                        verbose = FALSE,
                        country = "NO",
                        gisco_id = NULL) %>%
                mutate(union = rep("Norway", nrow(.))) %>%
                group_by(union) %>%
                summarise(., .groups = "drop") %>%
                st_transform(., crs = crs(pm25_data[[1]]))


# Mask
pm25_formatted <- pm25_data %>% 
                 map(., function(x){
                   
                   if(crs(x) == crs(no_mask)){
                     crop(x, no_mask, mask = TRUE)
                   }else{print("Error in CRS")}
                   
                 }) 

```


## 6. Export file
```{r}
#| eval: false
#| code-summary: "Export"

files_names <- str_extract(pm25_files, "pm25\\w*")

pm25_final_path <- map_vec(files_names, function(x){
  paste0(temp_dir,"/", x, "_NOmasked.tif")
})

names(pm25_final_path) <- names(pm25_data)

iwalk(pm25_formatted, function(layer, name) {
        writeRaster(x = layer,
                   filename = pm25_final_path[[name]],
                   overwrite = TRUE)})

```

## 7. Author contributions

<!-- Fill in CReDIT and MeRIT statments to clarify the contributions from individual authors. For MeRIT, see https://www.nature.com/articles/s41467-023-37039-1. For CRedit, see https://credit.niso.org/. You can start with the template structur below, and just delete the rows you dont use.  -->

**CRediT statement**

* Conceptualization: SC
* Methodology: SC & BC
* Data curation & formatting: SC
* Writing – original draft: SC
* Writing – review & editing: BC
* Document structure: SC & AK


**MeRIT attribution**

The structure of the document has been taken from ecRxiv's template for condition indicators, created by Anders Kolstad (AK). SC adapted it to the purpose of the project.

**Study design**
The study was designed by SC and BC.

**Data processing**
Raw data were downloaded by SC. They were curated and cleaned by SC and BC. Code was written by both SC and BC.


::: {.callout-tip collapse="true"}
## Session Info
<!-- You can leave this last part as it is, unless you have a good reason to alter it. -->

This information is mainly aimed at developers that need to recreate the R environment that produced the html.
Important note: this information is about the R session that was running when the html was rendered, and if some of the code chunks were not evaluated when the html was rendered, they may have been executed in a different environment.  

```{r sessionInfo}
#| echo: false
sessionInfo()
```

:::
