---
title: "Prepare Grunnkart data"
folderName: code # Leave to be filled out by ecRxiv during typesetting (name of the folder under 'indicators/'). 
# If it's a single indicator being reported, the folder name should be the same as the indicatorID. 
# If there are multiple ID's then the folder name could look something like XX_YYY_ZZZ-ÅÅÅ.
country: Norway
continent: Europe
yearAdded: 2026 # the year when the indicator was first submitted to ecRxiv
yearLastUpdate: 2026
status: incomplete # options are complete, incomplete, or deprecated
Version: "0.1.0" # See wiki page: https://github.com/NINAnor/ecRxiv/wiki#versioning
VersionComment: 'First version. The code has been adapted from https://github.com/NINAnor/ecRxiv/tree/main/indicators/NO_URGR_001-005. Template adapated from Anders Kolstad ecRxiv repository: https://github.com/NINAnor/ecRxiv.'
url: https://github.com/NINAnor/egd_data_preparation
# To be filled out by ecRxiv during the typesetting, when they will replace XX_YYYY_ZZ with the name of the root folder for the current indicator(s). It should be the same as the indicatorID, but if there are more indicator IDs presented in the same folder, the folder name might be different, e.g. XX_YYYY_AAA-BBB 
format: 
  html:
    embed-resources: true
    code-fold: true
    toc: true
    toc-title: Contents
    toc-depth: 3
    smooth-scroll: true
execute: 
  cache: true
author:
  - name: Sylvie Clappe             # Enter name
    email: sylvie.clappe@nina.no  # Enter email
    affiliations:
      - id: myID
        name: Norwegian Institute for Nature Research # Enter affiliations
  - name: Bálint Czúcz              
    email: balint.czucz@nina.no             #  Enter subsequent authors like this, or remove if not relevant
    affiliations:
      - ref: myID               # To reuse affiliations reference the id like this
AuthorList: Clappe, S., Czúcz, B.
date: July 13, 2026
bibliography: references.bib
callout-icon: false
lightbox: true
css: ../../../style.css
code-links:
      - text: Add a review
        icon: github
        href: https://github.com/NINAnor/ecRxiv
hide: TRUE # Change to FALSE
---

<!--# This is a template for how to document the indicator analyses. Make sure also to not change the order, or modify, the headers, unless you really need to. This is because it easier to read if all the indicators are presented using the same layout. If there is one header where you don't have anything to write, just leave the header as is, and don't write anything below it. If you are providing code, be careful to annotate and comment on every step in the analysis. Before starting it is recommended to fill in as much as you can in the metadata file. This file will populate the initial table in your output.-->

<!--# Load all you dependencies here -->

```{r setup}
#| include: false
#| cache: false
library(knitr)
library(tidyverse)
library(kableExtra)
library(here)
library(yaml)
library(tibble)
library(conflicted)
library(tidyverse)
library(dplyr)
library(sf)
library(carrier)
library(mirai)
library(duckdb)
library(duckspatial)

# weird workarounds for cache issues
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::lag)
conflicts_prefer(dplyr::select)
pull<-dplyr::pull

knitr::opts_chunk$set(echo = TRUE)
```

<!--# The following parts are autogenerated. Do not edit. -->


```{r source}
#| echo: false
#| cache: false

source(here::here("_common.R"))
```


```{r reviewers}
#| echo: false
#| results: asis

if (!is.null(yaml_data$codeReviewers)) {
  reviewers <- purrr::map_chr(yaml_data$codeReviewers, "name")
  
  htmltools::HTML(paste0(
    '<div id="title-block-header" class="quarto-title-block default">
       <div class="quarto-title-meta">
         <div>
           <div class="quarto-title-meta-heading">Reviewed by</div>
           <div class="quarto-title-meta-contents">',
    paste(reviewers, collapse = "; "),
    '</div>
         </div>
       </div>
     </div>',
    sep = ""
  ))
}
```

> **Recommended citation**: `r paste(auth, " ", year, ". ", "v. ", version, ": ", url, sep="")`



::: {.callout-tip collapse="true"}
## Metadata
```{r tbl-meta2}
#| tbl-cap: 'Indicator metadata'
#| echo: false
#| warning: false
meta |>
  dplyr::filter(Variable %in% c(
    "Indicator ID",
    "Indicator Name",
    "verbatimName",
    "Continent",
    "Country",
    "Year added", 
    "Last update",
    "Version",
    "Version comment")) |>   
  kable()

```
:::

::: {.callout-tip collapse="true"}

## Log

<!--# Update this logg with short messages for each update -->
- 13 July 2026 - First draft version
:::


<hr />

<!--# Document you work below.  -->

## 1. Summary
This code shows how to download and pre-process Grunnkart data to compile the URGR URAQ and TCCD indicators from ecRxiv: https://github.com/NINAnor/ecRxiv/tree/main/indicators/NO_URGR_001-005.


## 2. Data characteristics
Grunnkart is Norway's official *ecosystem type map* designed to support ecosystem accounting (@strand2024; @gk2). It is produced by NIBIO and SSB.

### 2.1 Spatial and temporal resolution and extent
Spatial resolution is not determined in the metadata, so is temporal resolution. The data are a composite of several datasets from different spatial and temporal resolution. For the purpose of our work, we will assume that the data are from 2025 as it is the date of its release.


### 2.2 Sampling design
The data are a composite of several datasets that have been compiled together. See NIBIO report @strand2024 for more information.

### 2.3 Original units
Original units are polygons.


### 2.4 Instructions for citing, using and accessing data
Data are available on Geonorge [here]https://kartkatalog.geonorge.no/metadata/nasjonalt-grunnkart-for-arealanalyse-testversjon-2/28c28e3a-d88f-4a34-8c60-5efe6d56a44d). Depending on the user's status, these data can be either freely available or available upon request only.


### 2.5 Additional comments about the dataset
Other metadata can be found [on Geonorge](https://kartkatalog.geonorge.no/metadata/nasjonalt-grunnkart-for-arealanalyse/28c28e3a-d88f-4a34-8c60-5efe6d56a44d?search=grunnkart)


## 3. Uncertainties
### 3.1 General uncertainties
These data have various sources of uncertainties but they could not be quantified and reported. These sources have not been discussed in the reports associated with the data but they are linked, among other things, to: the uncertainty of the different data sources used to compile the Grunnkart, the hierarchical choices made during the classification of ecosystem types, the crosswalk used between all the classifications used in the different datasets used to produce the Grunnkart.

### 3.2 Reported uncertainties
Uncertainties related to this dataset were not quantified or reported by NIBIO.


## 4. Load the data
The Grunnkart is a a big dataset and, for NINA, these data were only available upon request. This means that we could not read them using an API and they had to be downloaded and stored on NINA's internal servers directly. 

Prior testing revealed that `sf` package is less performant to correct geometrical issues than the more recent `duckspatial` package, which uses DuckDB. We run the first part of the code with `duckspatial` including loading the data and the cleaning of the geometries and slivers.

*Note: Most of the Grunnkart tiles are in EPSG:25832, two tiles are in EPSG:25833, one in EPSG:25835 and one in EPSG:4258. To reduce the number of operations, we choose to stay in EPSG:25832*

*Note: `duckspatial` creates a duckdb database to store the data. This can become heavy very easily so the solution is to either make sure to have space on the disk, or clean the intermediate product along the way*

The first step is to open a connection to DuckDB. We advise to do it on a temporary file that will be deleted later on. We limit the threads (i.e. cores) to 10 and the memory to 30GB.
```{r}
#| eval: false
#| code-summary: "Create a DuckDB connection"
conn_gk <- ddbs_create_conn(dbdir = "tempdir", threads = 10, memory_limit_gb = 30)

```

Now we can read the data and write them on our temporary DuckDB database. We re-project on the fly to EPSG:25832 if necessary. This process takes a bit of time due to the size of the data (~10mins).
```{r}
#| eval: false
#| code-summary: "Read the sf objects from gdb"

# Write data in duckdb connection and re-project if necessary 
# Path to data
gdb_folder <- "/data/R/GeoSpatialData/LandUse/Norway_Arealregnskap/Original/GrunnkartArealregnskap FGDB-format"

gdb_files <- list.files(gdb_folder, pattern = "_gdb\\.gdb$", 
                        full.names = TRUE)


# Main CRS - EPSG:25832
gk_crs <- st_read(gdb_files[1], 
                  query = "SELECT * FROM arealregnskap LIMIT 10") %>%
  st_crs()

# # Remove grunnkart_raw dataset if aleardy in my db connection
# dbExecute(conn_gk, "DROP TABLE IF EXISTS grunnkart_raw")

# Read
source(here::here("src", "functions", "ddbs_read_and_transform_gk.R"))

db_gk <- gdb_files %>%
  walk( ~ddbs_read_and_transform_gk(data_path_gdb = .,
                                    conn_name = conn_gk,
                                    crs_proj = gk_crs))

# Verify table is in my duckdb conn_gk database
ddbs_list_tables(conn_gk)

```

Now that we the raw data are written in the DuckDB database we established, we can move on to the cleaning step.


## 5. Data cleaning and formatting
*Note: DuckDB automatically paralellise code within the cores and memory limits we set.*

First we clean the slivers before re-projecting. We set polygons with an area less than 0.05m2 as slivers.
```{r}
#| eval: false
#| code-summary: "Clean slivers"

# recalculate areas and remove slivers
db_gk_sliver_cln <- ddbs_area("grunnkart_raw", new_column = "area_m2_proj", conn = conn_gk, overwrite = TRUE) %>%
                      filter(area_m2_proj > 0.05) 


# # Checks (if needed)
# gk <- ddbs_read_table(conn_gk, "grunnkart_raw")
# unique(gk$tile_name)
# gk_sliver_cln <- ddbs_collect(db_gk_sliver_cln, as = "sf")
# st_crs(gk_sliver_cln)
# nrow(gk) - nrow(gk_sliver_cln)

# Clean checks R Environment
rm(ddbs_read_and_transform_gk,
   gk,
   gk_sliver_cln)

```

Now we can fix geometry. 

```{r}
#| eval: false
#| code-summary: "Fix geometry issues"

# Fix issues and write the table in the database
ddbs_make_valid("grunnkart_raw",
                conn = conn_gk,
                "grunnkart_valid", # checkpoint - create a data table in duckdb database
                overwrite = TRUE)

# Verify table is in my duckdb conn_gk database
ddbs_list_tables(conn_gk)

# Checks 
valid_check <- ddbs_is_valid("grunnkart_valid",
                             conn = conn_gk) %>%
  filter(!is_valid) # same as is_valid == FALSE


empty_check <- ddbs_is_empty("grunnkart_valid",
                             conn = conn_gk) %>%
                  filter(is_empty == TRUE)

simple_check <- ddbs_is_simple("grunnkart_valid",
                               conn = conn_gk) %>%
                  filter(!is_simple)  # same as is_simple == FALSE



gk_valid_ck <- ddbs_collect(valid_check, as = "sf")
gk_empty_ck <- ddbs_collect(empty_check, as = "sf")
gk_simple_ck <- ddbs_collect(simple_check, as = "sf")

# Re-calculate areas
gk_valid <- ddbs_area("grunnkart_valid", new_column = "area_m2_valid", conn = conn_gk, overwrite = TRUE) %>%
                      select(tile_name, komid, okosystemtype_1, area_m2_valid, geo)

# Clean checks R Environment
rm(valid_check,
   empty_check,
   simple_check,
   gk_valid_ck,
   gk_empty_ck,
   gk_simple_ck)


```


Now we can export this intermediate product, to have one data saved in case the union crashes later. We also clean the duckdb database.
```{r}
#| eval: false
#| code-summary: "Fix geometry errors"

# Export the intermediate product
ddbs_write_dataset(gk_valid,
                   path = paste0(str_remove(tempdir(), "Rtmp\\w*"), "GK_raw/GK_valid_db_30JULY2026.gpkg"),
                   gdal_driver = "GPKG",
                   conn = conn_gk,
                   crs = "EPSG:25832",
                   overwrite = TRUE,
                   layer = "gk_valid")

# Clean duckdb databse
tables <- dbListTables(conn_gk)
tbl_list <- tables[!tables %in% "grunnkart_valid"]
            
for (tbl in tbl_list) {
  
  if(substr(tbl, 1, 9) == "temp_view"){
    
    dbExecute(conn_gk, sprintf("DROP VIEW IF EXISTS %s", dbQuoteIdentifier(conn_gk, tbl)))
    
  }else{
    
    dbExecute(conn_gk, sprintf("DROP TABLE IF EXISTS %s", dbQuoteIdentifier(conn_gk, tbl)))
  }
  
}

# Verify grunnkart_valid remains
ddbs_list_tables(conn_gk)

```



# BALINT ADDITIONAL CLEANING STEPS. 
I didn't do any union before as I thought you might need all the small polygons? I also left it as a list where each item is a tile. 

## 6. Data formatting
The goal is now to do a union of each ecosystem type across all the tiles (counties). To do that, we start by doing a union for each ecosystem type per tile. To avoid memory issues, we divide the tiles in three groups and perform the union on these three groups separately.
```{r}
#| eval: false
#| code-summary: "Dissolve ecosystem types into one single polygon"

# Merge ET level one per tiles 
gk_data_final <- gk_data_valid_fixed %>% 
                    map(., function(x){
                      x %>%
                        group_by(okosystemtype_1) %>%
                       summarise(geom = st_union(geom),
                                 area_m2_total = sum(area_m2_final), 
                                 .groups = "drop")
                    }) 


# Save intermediate product (if needed)
# gk_path_data_join <- paste0(str_remove(tempdir(), "Rtmp\\w*"), "GK_raw/GK_ETlevel1_union_tiles_21JULY2026.gpkg")
# 
# iwalk(gk_data_final, function(layer, name) {
#         st_write(obj = layer,
#                  dsn = gk_path_data_join,
#                  layer = name,   
#                  append = FALSE,    
#                  delete_layer = TRUE)})

# Read intermediate layer (if needed)
# gk_path_data_join <- paste0(str_remove(tempdir(), "Rtmp\\w*"), "GK_raw/GK_ETlevel1_union_tiles_21JULY2026.gpkg")
# 
# layer_names <- st_layers(gk_path_data_join)$name
# 
# gk_data_final <- map(layer_names, ~st_read(gk_path_data_join, layer = .x, quiet = TRUE))
# 
# names(gk_data_final) <- str_extract(gdb_files, "(?<=format/).{2}")

# Merge all tiles together and union ET level 1 at national level
gk_data_final <- gk_data_final %>% # overwrite data to spare memory
                   compact() %>%  
                   bind_rows() %>%
                   group_by(okosystemtype_1) %>%
                   summarise(geom = st_union(geom),
                             area_m2_total = sum(area_m2_total), 
                             area_m2_st = unclass(st_area(.)),
                             .groups = "drop")

gk_data_final_nat <- gk_data_final %>% # overwrite data to spare memory
    compact() %>%  
    bind_rows() %>%
    group_by(okosystemtype_1) %>%
    summarise(geom = st_union(geom),
              area_m2_total = sum(area_m2_total), 
              area_m2_st = unclass(st_area(.)),
              .groups = "drop")

# Check validity and CRS
final_validity <- which(st_is_valid(gk_data_final) == FALSE)
crs(gk_data_final[[1]])

# Export
gk_path_data_final <- paste0(str_remove(tempdir(), "Rtmp\\w*"), "GK_raw/GK_final_19JULY2026.gpkg")

st_write(gk_data_final, gk_path_data_final, layer = "ET level 1")


st_write(filter(gk_data_final, okosystemtype_1 == ""), 
         gk_data_final, 
         layer = "my_layer_name")

st_write(my_sf_object, gk_data_final, layer = "my_layer_name", append = TRUE)

# Clean R Environment
rm(list=setdiff(ls(), c("pull",
                        "gpkg_path",
                        "gpkg_path_proj",
                        "gpkg_path_sliver",
                        "gpkg_path_valid",
                        "gk_path_valid_fixed",
                        "gk_data_final")))


```


## 7. Export file
We export the data as a geopackage where each layer is an ecosystem type.
```{r export}
#| eval: false



```

## 8. References
<!--# You can add references manually or use a citation manager and add intext citations as with crossreferencing and hyperlinks. See https://quarto.org/docs/authoring/footnotes-and-citations.html -->


::: {#refs}
:::

## 9. Author contributions

<!-- Fill in CReDIT and MeRIT statments to clarify the contributions from individual authors. For MeRIT, see https://www.nature.com/articles/s41467-023-37039-1. For CRedit, see https://credit.niso.org/. You can start with the template structur below, and just delete the rows you dont use.  -->

**CRediT statement**

* Conceptualization: SC
* Methodology: SC & BC
* Data curation & formatting: SC
* Writing – original draft: SC
* Writing – review & editing: BC
* Document structure: SC & AK


**MeRIT attribution**

The structure of the document has been taken from ecRxiv's template for condition indicators, created by Anders Kolstad (AK). SC adapted it to the purpose of the project.

**Study design**
The study was designed by SC and BC.

**Data processing**
Raw data were downloaded by SC. They were curated and cleaned by SC and BC. Code was written by both SC and BC.


::: {.callout-tip collapse="true"}
## Session Info
<!-- You can leave this last part as it is, unless you have a good reason to alter it. -->

This information is mainly aimed at developers that need to recreate the R environment that produced the html.
Important note: this information is about the R session that was running when the html was rendered, and if some of the code chunks were not evaluated when the html was rendered, they may have been executed in a different environment.  

```{r sessionInfo}
#| echo: false
sessionInfo()
```

:::
