---
title: "Prepare Grunnkart data"
folderName: code # Leave to be filled out by ecRxiv during typesetting (name of the folder under 'indicators/'). 
# If it's a single indicator being reported, the folder name should be the same as the indicatorID. 
# If there are multiple ID's then the folder name could look something like XX_YYY_ZZZ-ÅÅÅ.
country: Norway
continent: Europe
yearAdded: 2026 # the year when the indicator was first submitted to ecRxiv
yearLastUpdate: 2026
status: incomplete # options are complete, incomplete, or deprecated
Version: "0.1.0" # See wiki page: https://github.com/NINAnor/ecRxiv/wiki#versioning
VersionComment: 'First version. The code has been adapted from https://github.com/NINAnor/ecRxiv/tree/main/indicators/NO_URGR_001-005. Template adapated from Anders Kolstad ecRxiv repository: https://github.com/NINAnor/ecRxiv.'
url: https://github.com/NINAnor/egd_data_preparation
# To be filled out by ecRxiv during the typesetting, when they will replace XX_YYYY_ZZ with the name of the root folder for the current indicator(s). It should be the same as the indicatorID, but if there are more indicator IDs presented in the same folder, the folder name might be different, e.g. XX_YYYY_AAA-BBB 
format: 
  html:
    embed-resources: true
    code-fold: true
    toc: true
    toc-title: Contents
    toc-depth: 3
    smooth-scroll: true
execute: 
  cache: true
author:
  - name: Sylvie Clappe             # Enter name
    email: sylvie.clappe@nina.no  # Enter email
    affiliations:
      - id: myID
        name: Norwegian Institute for Nature Research # Enter affiliations
  - name: Bálint Czúcz              
    email: balint.czucz@nina.no             #  Enter subsequent authors like this, or remove if not relevant
    affiliations:
      - ref: myID               # To reuse affiliations reference the id like this
AuthorList: Clappe, S., Czúcz, B.
date: July 13, 2026
bibliography: references.bib
callout-icon: false
lightbox: true
css: ../../../style.css
code-links:
      - text: Add a review
        icon: github
        href: https://github.com/NINAnor/ecRxiv
hide: TRUE # Change to FALSE
---

<!--# This is a template for how to document the indicator analyses. Make sure also to not change the order, or modify, the headers, unless you really need to. This is because it easier to read if all the indicators are presented using the same layout. If there is one header where you don't have anything to write, just leave the header as is, and don't write anything below it. If you are providing code, be careful to annotate and comment on every step in the analysis. Before starting it is recommended to fill in as much as you can in the metadata file. This file will populate the initial table in your output.-->

<!--# Load all you dependencies here -->

```{r setup}
#| include: false
#| cache: false
library(knitr)
library(tidyverse)
library(kableExtra)
library(here)
library(yaml)
library(tibble)
library(conflicted)
library(duckdb)
library(duckspatial)
library(sf)
library(dplyr)
library(stringr)

# weird workarounds for cache issues
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::lag)
conflicts_prefer(dplyr::select)
pull<-dplyr::pull

knitr::opts_chunk$set(echo = TRUE)
```

<!--# The following parts are autogenerated. Do not edit. -->


```{r source}
#| echo: false
#| cache: false

source(here::here("_common.R"))
```


```{r reviewers}
#| echo: false
#| results: asis

if (!is.null(yaml_data$codeReviewers)) {
  reviewers <- purrr::map_chr(yaml_data$codeReviewers, "name")
  
  htmltools::HTML(paste0(
    '<div id="title-block-header" class="quarto-title-block default">
       <div class="quarto-title-meta">
         <div>
           <div class="quarto-title-meta-heading">Reviewed by</div>
           <div class="quarto-title-meta-contents">',
    paste(reviewers, collapse = "; "),
    '</div>
         </div>
       </div>
     </div>',
    sep = ""
  ))
}
```

> **Recommended citation**: `r paste(auth, " ", year, ". ", "v. ", version, ": ", url, sep="")`



::: {.callout-tip collapse="true"}
## Metadata
```{r tbl-meta2}
#| tbl-cap: 'Indicator metadata'
#| echo: false
#| warning: false
meta |>
  dplyr::filter(Variable %in% c(
    "Indicator ID",
    "Indicator Name",
    "verbatimName",
    "Continent",
    "Country",
    "Year added", 
    "Last update",
    "Version",
    "Version comment")) |>   
  kable()

```
:::

::: {.callout-tip collapse="true"}

## Log

<!--# Update this logg with short messages for each update -->
- 13 July 2026 - First draft version
:::


<hr />

<!--# Document you work below.  -->

## 1. Summary
This code shows how to download and pre-process Grunnkart data to compile the URGR URAQ and TCCD indicators from ecRxiv: https://github.com/NINAnor/ecRxiv/tree/main/indicators/NO_URGR_001-005.


## 2. Data characteristics
Grunnkart is Norway's official *ecosystem type map* designed to support ecosystem accounting (@strand2024; @gk2). It is produced by NIBIO and SSB.

### 2.1 Spatial and temporal resolution and extent
Spatial resolution is not determined in the metadata, so is temporal resolution. The data are a composite of several datasets from different spatial and temporal resolution. For the purpose of our work, we will assume that the data are from 2025 as it is the date of its release.


### 2.2 Sampling design
The data are a composite of several datasets that have been compiled together. See NIBIO report @strand2024 for more information.

### 2.3 Original units
Original units are polygons.


### 2.4 Instructions for citing, using and accessing data
Data are available on Geonorge [here]https://kartkatalog.geonorge.no/metadata/nasjonalt-grunnkart-for-arealanalyse-testversjon-2/28c28e3a-d88f-4a34-8c60-5efe6d56a44d). Depending on the user's status, these data can be either freely available or available upon request only.


### 2.5 Additional comments about the dataset
Other metadata can be found [on Geonorge](https://kartkatalog.geonorge.no/metadata/nasjonalt-grunnkart-for-arealanalyse/28c28e3a-d88f-4a34-8c60-5efe6d56a44d?search=grunnkart)


## 3. Uncertainties
### 3.1 General uncertainties
These data have various sources of uncertainties but they could not be quantified and reported. These sources have not been discussed in the reports associated with the data but they are linked, among other things, to: the uncertainty of the different data sources used to compile the Grunnkart, the hierarchical choices made during the classification of ecosystem types, the crosswalk used between all the classifications used in the different datasets used to produce the Grunnkart.

### 3.2 Reported uncertainties
Uncertainties related to this dataset were not quantified or reported by NIBIO.


## 4. Load the data
The Grunnkart is a a big dataset and, for NINA, these data were only available upon request. This means that we could not read them using an API and they had to be downloaded and stored on NINA's internal servers directly. 

Prior testing revealed that `sf` package is less performant to correct geometrical issues than the more recent `duckspatial` package, which uses DuckDB. We run the first part of the code with `duckspatial` including loading the data and the cleaning of the geometries and slivers.

*Note: Most of the Grunnkart tiles are in EPSG:25832, two tiles are in EPSG:25833, one in EPSG:25835 and one in EPSG:4258. To reduce the number of operations, we choose to stay in EPSG:25832 and re-project everything in EPSG:25833, which is the default national projection, in section 7 when we export the data.*

*Note: `duckspatial` creates a duckdb database to store the data. This can become heavy very easily so the solution is to either make sure to have space on the disk, or clean the intermediate product along the way, which is what we do in this code.*

The first step is to open a connection to DuckDB. We advise to do it on a temporary file that will be deleted later on. We limit the threads (i.e. cores) to 10 and the memory to 30GB. This database can become quite heavy with data, that is why we clean it along the different cleaning steps to only keep the relevant data.
```{r}
#| eval: false
#| code-summary: "Create a DuckDB connection"
conn_gk <- ddbs_create_conn(dbdir = "tempdir", threads = 10, memory_limit_gb = 30)

```

Now we can read the data and write them on our temporary DuckDB database: one table per fylke. As we will do all the operations per fylke, we do not re-project and keep the original GK CRS for now. At the Export section, we will produce several geopackages with different projections. This process takes a bit of time due to the size of the data (~10mins).
```{r}
#| eval: false
#| code-summary: "Read the sf objects from gdb"

# Write data in duckdb connection and re-project if necessary 
# Path to data
gdb_folder <- "/data/R/GeoSpatialData/LandUse/Norway_Arealregnskap/Original/GrunnkartArealregnskap FGDB-format"

gdb_files <- list.files(gdb_folder, pattern = "_gdb\\.gdb$", 
                        full.names = TRUE)

# Read
source(here::here("src", "functions", "ddbs_read_and_transform_gk.R"))

db_gk <- gdb_files %>%
  walk( ~ddbs_read_and_transform_gk(data_path_gdb = .,
                                    conn_name = conn_gk))

# Verify tables are in my duckdb conn_gk database
ddbs_list_tables(conn_gk)

# Clean duckdb database
tables <- dbListTables(conn_gk) %>% 
            keep(., startsWith(., "temp_"))

for (tbl in tables) {

  if(substr(tbl, 1, 9) == "temp_view"){

    dbExecute(conn_gk, sprintf("DROP VIEW IF EXISTS %s", dbQuoteIdentifier(conn_gk, tbl)))

  }else{

    dbExecute(conn_gk, sprintf("DROP TABLE IF EXISTS %s", dbQuoteIdentifier(conn_gk, tbl)))

  }

}

```

Now that we the raw data are written in the DuckDB database we established, we can move on to the cleaning step.


## 5. Data cleaning and formatting
*Note: DuckDB automatically paralellise code within the cores and memory limits we set.*

First we clean the slivers before correcting for geometrical issues. We set polygons with an area less than 0.05m2 as slivers.
```{r}
#| eval: false
#| code-summary: "Clean slivers"

# recalculate areas and remove slivers
tables <- dbListTables(conn_gk) %>% 
            keep(., startsWith(., "grunnkart_raw"))

db_gk_sliver_cln <- map(tables,
                         ~ filter(as_duckspatial_df(.x, conn = conn_gk),
                                  areal_m2 > 0.05))

names(db_gk_sliver_cln) <- str_extract(tables, "(?<=_f).{2}")

# # Checks (if needed)
# gk <- ddbs_read_table(conn_gk, "grunnkart_raw")
# unique(gk$tile_name)
# gk_sliver_cln <- ddbs_collect(db_gk_sliver_cln, as = "sf")
# st_crs(gk_sliver_cln)
# nrow(gk) - nrow(gk_sliver_cln)

# Clean checks R Environment
rm(list = setdiff(ls(), c("conn_gk",
                          "pull",
                          "db_gk_sliver_cln")))

```

Now we can fix geometry. 
```{r}
#| eval: false
#| code-summary: "Fix geometry issues"

# Fix issues and write the table in the database
map(1:length(db_gk_sliver_cln), function(x){
  ddbs_make_valid(db_gk_sliver_cln[[x]],
                conn = conn_gk,
                name = paste0("grunnkart_valid_", names(db_gk_sliver_cln)[x]), 
                overwrite = TRUE)
}
)

# Verify table is in my duckdb conn_gk database
ddbs_list_tables(conn_gk)

# Checks - should all display 0 rows
tables <- dbListTables(conn_gk) %>%
            keep(., startsWith(., "grunnkart_valid_"))

for(tbl in tables){
  
  message("Processing: ", tbl)
  
  # Checks
  valid_check <- ddbs_is_valid(tbl, conn = conn_gk) %>%
                    filter(!is_valid) %>%
                    ddbs_collect() %>%
                    dim()

  empty_check <- ddbs_is_empty(tbl, conn = conn_gk) %>%
                    filter(is_empty == TRUE) %>%
                    ddbs_collect() %>%
                    dim()

  simple_check <- ddbs_is_simple(tbl, conn = conn_gk) %>%
                    filter(!is_simple) %>%
                    ddbs_collect() %>%
                    dim()
  
  # Print - should all have 0 rows
  print(valid_check)
  print(empty_check)
  print(simple_check)

  
}

# Clean duckdb database
tables <- dbListTables(conn_gk) %>% 
            keep(., startsWith(., "temp_") | startsWith(., "grunnkart_raw"))

for (tbl in tables) {

  if(substr(tbl, 1, 9) == "temp_view"){

    dbExecute(conn_gk, sprintf("DROP VIEW IF EXISTS %s", dbQuoteIdentifier(conn_gk, tbl)))

  }else{

    dbExecute(conn_gk, sprintf("DROP TABLE IF EXISTS %s", dbQuoteIdentifier(conn_gk, tbl)))

  }

}

# Clean checks R Environment
rm(list = setdiff(ls(), c("conn_gk",
                          "pull")))


```


Now we can export this intermediate product, to have one data saved in case the union crashes later.
```{r}
#| eval: false
#| code-summary: "Fix geometry errors"

## Export the intermediate product

# Path
path_gpkg <- paste0(str_remove(tempdir(), "Rtmp\\w*"), "GK_raw/GK_valid_14AUG2026.gpkg")

# Names of the fylke
tables <- dbListTables(conn_gk)
fylke_nb <- str_extract(tables, "(?<=valid_).{2}") %>%
              paste0("f", .)


walk(1:length(fylke_nb), function(x){
  
  as_duckspatial_df(tables[x], conn = conn_gk) %>%
    ddbs_collect() %>%
       st_write(.,
           path_gpkg,
           layer = fylke_nb[x],
           append = TRUE)
  }
  )

# Read intermediate layer (if necessary)
# path_gpkg <- paste0(str_remove(tempdir(), "Rtmp\\w*"), "GK_raw/GK_valid_14AUG2026.gpkg")
# 
# layers <- st_layers(path_gpkg)
# layers <- layers$name
# 
# walk(layers, function(x){
#   
#   dbExecute(conn_gk, 
#       sprintf("
#         CREATE TABLE '%s' AS
#         SELECT *
#         FROM ST_Read('%s', layer = '%s')
#       ", 
# paste0("grunnkart_valid_", layers[x])),
# path_gpkg, layers[x])
#   
# })

# Clean checks R Environment
rm(list = setdiff(ls(), c("conn_gk",
                          "pull")))

```


## 6. Data formatting
The goal is now to do a union of each ecosystem type per tile (counties). To avoid memory issues, we have to do block processing and do the union county by county. It didn't make much sense to union at national level (across tiles) notably because, when we started to union the ecosystem types across tiles, the number of polygons did not change that much. This is because the only places where the polygons might union are at the border of the tiles. Moreover, for URAQ and URGR, intersections are needed with Local Administrative Units, so to decrease the computing time, keeping the information on the county is very useful as we can then just download the county(ies) of relevance instead of the whole dataset.

Doing a union per tile and per ecosystem type will resulst in each tile having merged polygons per ecosystem types present. In the code below, we use the function `ddbs_union_gk()` which we created to do several steps at once: (i) union per tile and ecosystem types; (ii) cast to polygon geometry only; (iii) make the geometry valid. We save the results in a geopackage along the way in case of unexpected crash.
**Important: step (ii) is very important, keeping complex geometries such as multipolygons and geometry collections hinders the spatial calculations as they are difficult to deal with for GDAl and GEOS. That is why, it is essential to always come back to polygons**.

*Note: DuckDB automatically parallelises processes, hence it is very important to set cores and memory limits in the duckdb connection! With 10 cores and a memory limit of 30GB, it takes about 1h-2h for the union and cleaning to run on one dataset. So in total, the code below should take about 4-5h (patience is a virtue!). We save each intermediate product as a geopackage in case one of the unions crash.*

```{r}
#| eval: false
#| code-summary: "Union per tile and ecosystem type - 3 subdatasets"

# Function to perform union and return a clean layer
source(here::here("src", "functions", "ddbs_union_gk.R"))

# Path to geopackage to save unions along the way
path_gpkg <- paste0(str_remove(tempdir(), "Rtmp\\w*"), "GK_raw/GK_union_14AUG2026.gpkg")

# List of tables and fylke number
tables <- dbListTables(conn_gk)
fylke_nb <- str_extract(tables, "(?<=valid_).{2}") %>%
              paste0("f", .)

walk(1:length(tables), function(x){
  
  message("Processing: ", tables[x])
  
  # Do union
  as_duckspatial_df(tables[x], conn = conn_gk) %>%
  
  ddbs_union_gk(duckspatial_obj = .,
                union_col = c("okosystemtype_1"),
                conn_name = conn_gk, 
                output_name = paste0("grunnkart_union_", fylke_nb[x]))

  # Export as a geopackage
  as_duckspatial_df(paste0("grunnkart_union_", fylke_nb[x]), conn = conn_gk) %>% 
        ddbs_collect(as = "sf") %>%
          st_write(dsn = path_gpkg,
                   layer = fylke_nb[x],
                   append = TRUE)
  
}

)

# Read intermediate layer (if necessary)
path_gpkg <- paste0(str_remove(tempdir(), "Rtmp\\w*"), "GK_raw/GK_union_14AUG2026.gpkg")
layers <- st_layers(path_gpkg)
layers <- layers$name

walk(1: length(layers), function(x){

  dbExecute(conn_gk,
      sprintf("
        CREATE OR REPLACE TABLE '%s' AS
        SELECT *
        FROM ST_Read('%s', layer = '%s')
      ",
paste0("grunnkart_union_", layers[x]),
path_gpkg, 
layers[x]))

})

# Check geometry
tables <- dbListTables(conn_gk) %>%
            keep(., startsWith(., "grunnkart_union_"))

for(tbl in tables){
  
  message("Processing: ", tbl)
  
  # Checks
  valid_check <- ddbs_is_valid(tbl, conn = conn_gk) %>%
                    filter(!is_valid) %>%
                    ddbs_collect() %>%
                    dim()

  empty_check <- ddbs_is_empty(tbl, conn = conn_gk) %>%
                    filter(is_empty == TRUE) %>%
                    ddbs_collect() %>%
                    dim()

  simple_check <- ddbs_is_simple(tbl, conn = conn_gk) %>%
                    filter(!is_simple) %>%
                    ddbs_collect() %>%
                    dim()
  
  # Print - should all have 0 rows
  print(valid_check)
  print(empty_check)
  print(simple_check)

  
}

# Clean duckdb databse
tables <- dbListTables(conn_gk) %>% 
            keep(., startsWith(., "temp_") | startsWith(., "grunnkart_valid"))

for (tbl in tables) {

  if(substr(tbl, 1, 9) == "temp_view"){

    dbExecute(conn_gk, sprintf("DROP VIEW IF EXISTS %s", dbQuoteIdentifier(conn_gk, tbl)))

  }else{

    dbExecute(conn_gk, sprintf("DROP TABLE IF EXISTS %s", dbQuoteIdentifier(conn_gk, tbl)))
  }

}

# Verify grunnkart_union remains
ddbs_list_tables(conn_gk)

# Clean checks R Environment
rm(list = setdiff(ls(), c("conn_gk",
                          "pull")))

```


## 7. Export file

### 7.1 Export extent-relevant data
Exports were done in two projections: ESPG 25833 and 25832. For the geopackages exported per county (fylke), we also exported with the projection inherited from the GK. We re-calculate the area before exporting. *Note: DuckDB's function to calculate the area popped a warning when calculating the area using the CRS inherited from the original GK. One of the projection used created some mathematical inconsistencies. Areas should thus be used with care for when the original GK projection is used.*

#### 7.1.1 Export files per county
```{r export}
#| eval: false
#| code-summary: "Export in original county projection"

# Path
path_gpkg <- paste0(str_remove(tempdir(), "Rtmp\\w*"), "GK_raw/GK_final_original_crs_")

# List of tables in duckdb database
tables <- dbListTables(conn_gk) %>% 
            keep(., startsWith(., "grunnkart_union"))

# Fylke number
fylke_nb <- str_extract(tables, "(?<=union_).{3}")

walk(1: length(fylke_nb), function(x){
  
  gk_fyk_tbl<-  as_duckspatial_df(tables[x], conn_gk) %>%
                ddbs_area(new_column = "area_m2") %>% 
                ddbs_collect() %>%
                select(!fid) %>% # if from intermediate layer!
                rename(fylke_code = tile_name) %>%
                rename(okosystemtype1_name = okosystemtype_1) %>%
                mutate(okosystemtype1_code = if_else(okosystemtype1_name == "Settlements and other artificial areas", "e01", 
                                                     ifelse(okosystemtype1_name == "Cropland", "e02",
                                                            ifelse(okosystemtype1_name == "Grassland", "e03",
                                                                   ifelse(okosystemtype1_name == "Forest and woodlands", "e04", 
                                                                          ifelse(okosystemtype1_name == "Heathland and shrub", "e05",
                                                                                 ifelse(okosystemtype1_name == "Sparsely vegetated ecosystems", "e06",
                                                                                        ifelse(okosystemtype1_name == "Inland wetlands", "e07",
                                                                                        ifelse(okosystemtype1_name == "Rivers and canals", "e08",
                                                                                        ifelse(okosystemtype1_name == "Lakes and reservoirs", "e09",
                                                                                        ifelse(okosystemtype1_name == "Marine inlets and transitional waters", "e10",
                                                                                        ifelse(okosystemtype1_name == "Coastal beaches, dunes and wetlands", "e11",
                                                                                        ifelse(okosystemtype1_name == "Marine ecosystems", "e12",
                                                                                        ifelse("ERROR"))))))))))))))
    
  
  st_write(gk_fyk_tbl,
           paste0(path_gpkg, fylke_nb[x], ".gpkg"),
           append = FALSE)
  
})

# Clean duckdb databse
tables <- dbListTables(conn_gk) %>% 
            keep(., startsWith(., "temp_"))

for (tbl in tables) {

  if(substr(tbl, 1, 9) == "temp_view"){

    dbExecute(conn_gk, sprintf("DROP VIEW IF EXISTS %s", dbQuoteIdentifier(conn_gk, tbl)))

  }else{

    dbExecute(conn_gk, sprintf("DROP TABLE IF EXISTS %s", dbQuoteIdentifier(conn_gk, tbl)))
  }

}

# Clean checks R Environment
rm(list = setdiff(ls(), c("conn_gk",
                          "pull")))
  

```

```{r export}
#| eval: false
#| code-summary: "Export in EPSG:25832"

# Path
path_gpkg <- paste0(str_remove(tempdir(), "Rtmp\\w*"), "GK_raw/GK_final_UTM32_")

# List of tables in duckdb database
tables <- dbListTables(conn_gk) %>% 
            keep(., startsWith(., "grunnkart_union"))

# Fylke number
fylke_nb <- str_extract(tables, "(?<=union_).{3}")

# Create geopackage
walk(1: length(fylke_nb), function(x){
  
  gk_fyk_tbl<-  ddbs_transform(x = tables[x], y = "EPSG:25832", conn = conn_gk) %>%
                ddbs_make_valid() %>%
                ddbs_area(new_column = "area_m2") %>%
                ddbs_collect() %>%
                select(!fid) %>% # if from intermediate layer!
                rename(fylke_code = tile_name) %>%
                rename(okosystemtype1_name = okosystemtype_1) %>%
                mutate(okosystemtype1_code = if_else(okosystemtype1_name == "Settlements and other artificial areas", "e01", 
                                                     ifelse(okosystemtype1_name == "Cropland", "e02",
                                                            ifelse(okosystemtype1_name == "Grassland", "e03",
                                                                   ifelse(okosystemtype1_name == "Forest and woodlands", "e04", 
                                                                          ifelse(okosystemtype1_name == "Heathland and shrub", "e05",
                                                                                 ifelse(okosystemtype1_name == "Sparsely vegetated ecosystems", "e06",
                                                                                        ifelse(okosystemtype1_name == "Inland wetlands", "e07",
                                                                                        ifelse(okosystemtype1_name == "Rivers and canals", "e08",
                                                                                        ifelse(okosystemtype1_name == "Lakes and reservoirs", "e09",
                                                                                        ifelse(okosystemtype1_name == "Marine inlets and transitional waters", "e10",
                                                                                        ifelse(okosystemtype1_name == "Coastal beaches, dunes and wetlands", "e11",
                                                                                        ifelse(okosystemtype1_name == "Marine ecosystems", "e12",
                                                                                        ifelse("ERROR"))))))))))))))
    
  
  st_write(gk_fyk_tbl,
           paste0(path_gpkg, fylke_nb[x], ".gpkg"),
           append = FALSE)
  
})

# Clean duckdb databse
tables <- dbListTables(conn_gk) %>% 
            keep(., startsWith(., "temp_"))

for (tbl in tables) {

  if(substr(tbl, 1, 9) == "temp_view"){

    dbExecute(conn_gk, sprintf("DROP VIEW IF EXISTS %s", dbQuoteIdentifier(conn_gk, tbl)))

  }else{

    dbExecute(conn_gk, sprintf("DROP TABLE IF EXISTS %s", dbQuoteIdentifier(conn_gk, tbl)))
  }

}

# Clean checks R Environment
rm(list = setdiff(ls(), c("conn_gk",
                          "pull")))

```

```{r export}
#| eval: false
#| code-summary: "Export in EPSG:25833"

## Path
path_gpkg <- paste0(str_remove(tempdir(), "Rtmp\\w*"), "GK_raw/GK_final_UTM33_")

# List of tables in duckdb database
tables <- dbListTables(conn_gk) %>% 
            keep(., startsWith(., "grunnkart_union"))

# Fylke number
fylke_nb <- str_extract(tables, "(?<=union_).{3}")

# Create geopackage
walk(1: length(fylke_nb), function(x){
  
  gk_fyk_tbl<-  ddbs_transform(x = tables[x], y = "EPSG:25833", conn = conn_gk) %>%
                ddbs_make_valid() %>%
                ddbs_area(new_column = "area_m2") %>%
                ddbs_collect() %>%
                select(!fid) %>% # if from intermediate layer!
                rename(fylke_code = tile_name) %>%
                rename(okosystemtype1_name = okosystemtype_1) %>%
                mutate(okosystemtype1_code = if_else(okosystemtype1_name == "Settlements and other artificial areas", "e01", 
                                                     ifelse(okosystemtype1_name == "Cropland", "e02",
                                                            ifelse(okosystemtype1_name == "Grassland", "e03",
                                                                   ifelse(okosystemtype1_name == "Forest and woodlands", "e04", 
                                                                          ifelse(okosystemtype1_name == "Heathland and shrub", "e05",
                                                                                 ifelse(okosystemtype1_name == "Sparsely vegetated ecosystems", "e06",
                                                                                        ifelse(okosystemtype1_name == "Inland wetlands", "e07",
                                                                                        ifelse(okosystemtype1_name == "Rivers and canals", "e08",
                                                                                        ifelse(okosystemtype1_name == "Lakes and reservoirs", "e09",
                                                                                        ifelse(okosystemtype1_name == "Marine inlets and transitional waters", "e10",
                                                                                        ifelse(okosystemtype1_name == "Coastal beaches, dunes and wetlands", "e11",
                                                                                        ifelse(okosystemtype1_name == "Marine ecosystems", "e12",
                                                                                        ifelse("ERROR"))))))))))))))
    
  
  st_write(gk_fyk_tbl,
           paste0(path_gpkg, fylke_nb[x], ".gpkg"),
           append = FALSE)
  
})

# Clean duckdb databse
tables <- dbListTables(conn_gk) %>% 
            keep(., startsWith(., "temp_"))

for (tbl in tables) {

  if(substr(tbl, 1, 9) == "temp_view"){

    dbExecute(conn_gk, sprintf("DROP VIEW IF EXISTS %s", dbQuoteIdentifier(conn_gk, tbl)))

  }else{

    dbExecute(conn_gk, sprintf("DROP TABLE IF EXISTS %s", dbQuoteIdentifier(conn_gk, tbl)))
  }

}

# Clean checks R Environment
rm(list = setdiff(ls(), c("conn_gk",
                          "pull")))

```


#### 7.1.2 Export files per ecosystem type
Now, we are just going to export one geopackage per ET, both in EPSG 25833 and 25832.

```{r export}
#| eval: false
#| code-summary: "Export in EPSG:25833"

## Path
path_gpkg <- paste0(str_remove(tempdir(), "Rtmp\\w*"), "GK_raw/GK_final_ecosystemtype_UTM33_")

# List of tables in duckdb database
tables <- dbListTables(conn_gk) %>% 
            keep(., startsWith(., "grunnkart_union"))

# Fylke number
fylke_nb <- str_extract(tables, "(?<=union_).{3}")


# Re-project fylke and union them
walk(1 : length(fylke_nb), function(x){
  
  message("Processing fylke: ", fylke_nb[x])
  
  
  # Project and compile area
  ddbs_sd <- ddbs_transform(x = tables[x], y = "EPSG:25833", conn = conn_gk) %>%
                ddbs_make_valid() %>%
                ddbs_area(new_column = "area_m2")
  
  # Write the table in the duckdb database
  if (!DBI::dbExistsTable(conn_gk, "grunnkart_all_33")) {

    # First tile: create the table
    ddbs_sd %>%
      ddbs_write_table(conn = conn_gk, name = "grunnkart_all_33")

  } else {

    # Subsequent tiles: append
    rows_append(tbl(conn_gk, "grunnkart_all_33"),
                ddbs_sd,
                in_place = TRUE)
  }
  
})

# Check table is in the duckdb database
ddbs_list_tables(conn_gk)

# List of ETs
et_list <- dbGetQuery(conn_gk, "SELECT DISTINCT okosystemtype_1 FROM grunnkart_all_33")[[1]]

# Create geopackage
walk(1: length(et_list), function(x){
  
  message("Processing ET: ", et_list[x])
  
  gk_fyk_tbl<-  as_duckspatial_df("grunnkart_all_33", conn = conn_gk) %>%
                filter(okosystemtype_1 == !!et_list[[x]]) %>%
                ddbs_collect() %>%
                select(!fid) %>% # if from intermediate layer!
                rename(fylke_code = tile_name) %>%
                rename(okosystemtype1_name = okosystemtype_1) %>%
                mutate(okosystemtype1_code = if_else(okosystemtype1_name == "Settlements and other artificial areas", "e01", 
                                                     ifelse(okosystemtype1_name == "Cropland", "e02",
                                                            ifelse(okosystemtype1_name == "Grassland", "e03",
                                                                   ifelse(okosystemtype1_name == "Forest and woodlands", "e04", 
                                                                          ifelse(okosystemtype1_name == "Heathland and shrub", "e05",
                                                                                 ifelse(okosystemtype1_name == "Sparsely vegetated ecosystems", "e06",
                                                                                        ifelse(okosystemtype1_name == "Inland wetlands", "e07",
                                                                                        ifelse(okosystemtype1_name == "Rivers and canals", "e08",
                                                                                        ifelse(okosystemtype1_name == "Lakes and reservoirs", "e09",
                                                                                        ifelse(okosystemtype1_name == "Marine inlets and transitional waters", "e10",
                                                                                        ifelse(okosystemtype1_name == "Coastal beaches, dunes and wetlands", "e11",
                                                                                        ifelse(okosystemtype1_name == "Marine ecosystems", "e12",
                                                                                        ifelse("ERROR"))))))))))))))
    
  
  st_write(gk_fyk_tbl,
           paste0(path_gpkg, unique(gk_fyk_tbl$okosystemtype1_code), ".gpkg"),
           append = FALSE)
  
})

# Clean duckdb databse
tables <- dbListTables(conn_gk) %>% 
            keep(., startsWith(., "temp_") | startsWith(., "grunnkart_all"))

for (tbl in tables) {

  if(substr(tbl, 1, 9) == "temp_view"){

    dbExecute(conn_gk, sprintf("DROP VIEW IF EXISTS %s", dbQuoteIdentifier(conn_gk, tbl)))

  }else{

    dbExecute(conn_gk, sprintf("DROP TABLE IF EXISTS %s", dbQuoteIdentifier(conn_gk, tbl)))
  }

}

# Clean checks R Environment
rm(list = setdiff(ls(), c("conn_gk",
                          "pull")))

```


```{r export}
#| eval: false
#| code-summary: "Export in EPSG:25832"

## Path
path_gpkg <- paste0(str_remove(tempdir(), "Rtmp\\w*"), "GK_raw/GK_final_ecosystemtype_UTM32_")

# List of tables in duckdb database
tables <- dbListTables(conn_gk) %>% 
            keep(., startsWith(., "grunnkart_union"))

# Fylke number
fylke_nb <- str_extract(tables, "(?<=union_).{3}")


# Re-project fylke and union them
walk(1 : length(fylke_nb), function(x){
  
  message("Processing fylke: ", fylke_nb[x])
  
  # Project and compile area
  ddbs_sd <- ddbs_transform(x = tables[x], y = "EPSG:25832", conn = conn_gk) %>%
                ddbs_make_valid() %>%
                ddbs_area(new_column = "area_m2")
  
  # Write the table in the duckdb database
  if (!DBI::dbExistsTable(conn_gk, "grunnkart_all_32")) {

    # First tile: create the table
    ddbs_sd %>%
      ddbs_write_table(conn = conn_gk, name = "grunnkart_all_32")

  } else {

    # Subsequent tiles: append
    rows_append(tbl(conn_gk, "grunnkart_all_32"),
                ddbs_sd,
                in_place = TRUE)
  }
  
})

# Check table is in the duckdb database
ddbs_list_tables(conn_gk)

# List of ETs
et_list <- dbGetQuery(conn_gk, "SELECT DISTINCT okosystemtype_1 FROM grunnkart_all_32")[[1]]

# Create geopackage
walk(1: length(et_list), function(x){
  
  message("Processing ET: ", et_list[x])
  
  gk_fyk_tbl<-  as_duckspatial_df("grunnkart_all_32", conn = conn_gk) %>%
                filter(okosystemtype_1 == !!et_list[[x]]) %>%
                ddbs_collect() %>%
                select(!fid) %>% # if from intermediate layer!
                rename(fylke_code = tile_name) %>%
                rename(okosystemtype1_name = okosystemtype_1) %>%
                mutate(okosystemtype1_code = if_else(okosystemtype1_name == "Settlements and other artificial areas", "e01", 
                                                     ifelse(okosystemtype1_name == "Cropland", "e02",
                                                            ifelse(okosystemtype1_name == "Grassland", "e03",
                                                                   ifelse(okosystemtype1_name == "Forest and woodlands", "e04", 
                                                                          ifelse(okosystemtype1_name == "Heathland and shrub", "e05",
                                                                                 ifelse(okosystemtype1_name == "Sparsely vegetated ecosystems", "e06",
                                                                                        ifelse(okosystemtype1_name == "Inland wetlands", "e07",
                                                                                        ifelse(okosystemtype1_name == "Rivers and canals", "e08",
                                                                                        ifelse(okosystemtype1_name == "Lakes and reservoirs", "e09",
                                                                                        ifelse(okosystemtype1_name == "Marine inlets and transitional waters", "e10",
                                                                                        ifelse(okosystemtype1_name == "Coastal beaches, dunes and wetlands", "e11",
                                                                                        ifelse(okosystemtype1_name == "Marine ecosystems", "e12",
                                                                                        ifelse("ERROR"))))))))))))))
    
  
  st_write(gk_fyk_tbl,
           paste0(path_gpkg, unique(gk_fyk_tbl$okosystemtype1_code), ".gpkg"),
           append = FALSE)
  
})

# Close duckdb database
ddbs_stop_conn(conn_gk)

# CleanR Environment
rm(list = ls())

```



# 8. References
<!--# You can add references manually or use a citation manager and add intext citations as with crossreferencing and hyperlinks. See https://quarto.org/docs/authoring/footnotes-and-citations.html -->


::: {#refs}
:::

## 9. Author contributions

<!-- Fill in CReDIT and MeRIT statments to clarify the contributions from individual authors. For MeRIT, see https://www.nature.com/articles/s41467-023-37039-1. For CRedit, see https://credit.niso.org/. You can start with the template structur below, and just delete the rows you dont use.  -->

**CRediT statement**

* Conceptualization: SC
* Methodology: SC & BC
* Data curation & formatting: SC
* Writing – original draft: SC
* Writing – review & editing: BC
* Document structure: SC & AK


**MeRIT attribution**

The structure of the document has been taken from ecRxiv's template for condition indicators, created by Anders Kolstad (AK). SC adapted it to the purpose of the project.

**Study design**
The study was designed by SC and BC.

**Data processing**
Raw data were downloaded by SC. They were curated and cleaned by SC and BC. Code was written by both SC and BC.


::: {.callout-tip collapse="true"}
## Session Info
<!-- You can leave this last part as it is, unless you have a good reason to alter it. -->

This information is mainly aimed at developers that need to recreate the R environment that produced the html.
Important note: this information is about the R session that was running when the html was rendered, and if some of the code chunks were not evaluated when the html was rendered, they may have been executed in a different environment.  

```{r sessionInfo}
#| echo: false
sessionInfo()
```

:::
